### 2023-12-17 Article.tex
init

\RequirePackage[2020-02-02]{latexrelease}
\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algcompatible}

% Please give the surname of the lead author for the running footer
\leadauthor{Šmelko}

\begin{document}

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
\shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\Letter]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5,\Letter]{Vincent Noël}

\affil[1]{Department of Software Engineering, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
\maketitle

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Computational models in systems biology are becoming more important with the advancement of experimental techniques to query the mechanistic details responsible for leading to phenotypes of interest. In particular, Boolean models are well fit to describe the complexity of signaling networks while being simple enough to scale to a very large number of components. With the advance of Boolean model inference techniques, the field is transforming from an artisanal way of building models of moderate size to a more automatized one, leading to very large models. In this context, adapting the simulation software for such increases in complexity is crucial. 
We present two new developments in the continuous time Boolean simulators: MaBoSS.MPI, a parallel implementation of MaBoSS which can exploit the computational power of very large CPU clusters, and MaBoSS.GPU, which can use GPU accelerators to perform these simulations. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}

MaBoSS \cite{stoll2012continuous, stoll2017maboss}\\
Why Boolean models are important\\
Why are they getting larger\\
Why are they need to be adapted for HPC environment\\

\section*{Background}

In this section, we describe the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.

MaBoSS algorithm takes the Boolean approach for modeling networks which is defined by a Boolean model. A model consists of $n$ nodes which can be in one of two Boolean values: active ($1$) or inactive ($0$). The \emph{state} of the whole modeled system is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$ and it holds that $|\mathcal{S}| = 2^n$.

The model interactions are described as transitions between two states. A single state can have multiple transitions to other states; to encode the notion of prioritizing a specific state transition over others, the transitions are assigned a non-negative real value. Simply, a Boolean network generalizes to a directed weighted graph $G = (\mathcal{S}, \rho)$, where $\rho: \mathcal{S} \times \mathcal{S} \rightarrow [0, \infty)$ is a transition function generating \emph{transition rates}. For convenience, it holds that:
\begin{equation}
    \rho(S, S') = 0 \iff \text{there is no transition from}\ S\ \text{to}\ S'
\end{equation}

MaBoSS simulates \emph{asynchronous update strategy}, further restricting the transitions between states. In this strategy, only a single node can change its value in a single transition; i.e., for a transition from $S$ to $S'$, there exists a single node $i$ such that $S_i \neq S'_i$ and for all $j \neq i$, $S_j = S'_j$. Consequently, a state $S$ can have at most $n$ possible transitions.

For each node, a \emph{Boolean logic} $\mathcal{B}: \mathcal{S} \rightarrow [0, \infty)$ is specified. The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$ (where only $S_i$ and $S'_i$ differ). If $\mathcal{B}_i(S) = 0$, then the transition is not allowed. Otherwise, the transition is allowed, and the result value determines the transition rate.

This formalization can also be understood as a formal definition of the Continuous-time Markov process. MaBoSS algorithm simulates this process by producing stochastic \emph{trajectories}. A trajectory is a sequence of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends when a state with no outgoing transitions or a maximum simulation time is reached.

Given a state $S$ and time $t$, the pseudocode for a single iteration of a trajectory simulation is as follows:
\begin{algorithmic}[1]
\State Compute transition rates $\rho_1 = \mathcal{B}_1(S), \dots, \rho_n = \mathcal{B}_n(S)$
\State Sample a random number $r$ from $[0, \sum_{i=1}^n \rho_i)$
\State Find $i$ such that $\sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$
\State Compute $S'$ from $S$ by flipping the $i$-th bit
\State Sample a random number $u$ from $[0, 1]$
\State Compute $\delta t = -(\ln u) / \sum_{i=1}^n \rho_i$
\State Set $S = S'$ and $t = t + \delta t$
\end{algorithmic}

Multiple trajectories are generated to get a good representation of the Markov process. Then, the statistics of the trajectories are computed. These are:

\begin{enumerate}
    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size $\Delta t$. For each window, the probability of each state is computed as the duration spent in the state divided by the window size. The probabilities of the corresponding windows are then averaged across all subtrajectories.
    \item \emph{Final states} --- The final state of a trajectory is the last state of the trajectory. The sampled states from the trajectories are used to compute the final state distribution.
    \item \emph{Fixed states} --- A state is fixed if it has no outgoing transitions. The sampled states from the trajectories are used to compute the fixed state distribution.
\end{enumerate}

Notably, nodes can be marked as \emph{internal}. The internal nodes are not considered in the statistics computation of time window probabilities and final states. More specifically, before a trajectory is used for statistics computation, the internal nodes are removed from each state in the trajectory.
This feature is helpful for the models where some nodes are used only for simulation purposes and/or are irrelevant to the final results.

\section*{Complexity evaluation and parallelization opportunities of MaBoSS}

MaBoSS algorithm is, in many aspects, very suitable for highly parallel processing. The simulation across multiple trajectories is embarrassingly parallel because the trajectories are independent of each other. Although the statistics computation is slightly more complicated because the intermediate results from separate trajectories need to be reduced to compute the final statistics, the topic of parallel reduction is a well-researched problem with many efficient solutions.

In this section, we detail the time complexity of the algorithm steps and discuss the parallelization opportunities.

\subsection*{Simulation complexity and parallelization}

Let us start by analyzing the time complexity of one iteration step of the simulation. Let us have a model with $n$ nodes. If we assume that one Boolean logic formula can be evaluated in $\mathcal{O}(n)$, then the computation of transition rates can be done in $\mathcal{O}(n^2)$. The selection of the flipping bit is $\mathcal{O}(n)$ and the rest of the iteration is $\mathcal{O}(1)$. Therefore, the total time complexity of one iteration is $\mathcal{O}(n^2)$. If we simulate $c$ trajectories with an upper bound of trajectory length $u$, the total simulation time is $\mathcal{O}(c \cdot u \cdot n^2)$.

Using an idealized parallel execution machine (PRAM model with infinite parallelism), we can optimize the algorithm in the following ways: 
\begin{itemize}
    \item All trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$.
    \item A single trajectory iteration can be parallelized. With $n$ processors, the computation of transition rates is achievable in $\mathcal{O}(n)$ time, where each processor would evaluate one Boolean logic formula, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using parallel prefix sum. This sums up to $\mathcal{O}(n)$ time for a single iteration. 
\end{itemize}

Overall, using a perfect parallel machine with $c \cdot n$ processors, the simulation time can be reduced to $\mathcal{O}(u \cdot n)$. There is a clear dependency between the trajectory iterations, so $\mathcal{O}(u)$ steps must be performed serially. Realistically, parallelizing an iteration step may prove beneficial only for rather large models such that the scheduling overhead and communications cost are lower than the achieved parallelism.

Also, since the trajectory iteration is the main `hot loop' of the whole algorithm (it is executed $c \cdot u$ times), optimizing it as much as possible is imperative. The penalty for the processor handling a cache miss inside each loop can result in the orders-of-magnitude slowdown. This puts a big pressure on the choice of the data representation for the state $S$ and Boolean logic formulas. We discuss this topic in the next section.

\subsection*{Statistics computation complexity and parallelization}

For the sake of brevity, let us discuss only the time window probabilities computation. The final and fixed states computation performs the same operation but less often because the final and fixed states are computed only from the last state of each trajectory.

The first part of the computation is updating a data structure that holds the durations of states visited in a trajectory. With the choice of a hash map as the data structure, the update can be done in $\mathcal{O}(1)$ time. In the second part, after all trajectories are simulated, the intermediate data structures from the trajectories must be merged into a final one. This can be done in $\mathcal{O}(\log{c} \cdot m)$ time, assuming a serial merging cost of $\mathcal{O}(m)$.

Notably, if the number of states is small enough, or the nodes are dominated by many internal nodes, the hash map can be materialized as a fixed-size array (and be interpreted as a histogram). This would eliminate any memory allocations and the key lookup would be reduced to a simple index function. This optimization is discussed in the next section.

In the following section, we describe the implementation of MaBoSS.MPI and MaBoSS.GPU. We use this discussion to detail the specific optimization choices in the implementation of the algorithms.

\section*{Implementation}
\subsection*{MaBoSS.MPI}

\subsection*{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and binary expression trees, that are not well suited for GPU hardware. In this section, we describe a CUDA GPU implementation of MaBoSS, and we detail the optimizations used to enable higher performance.

The MaBoSS algorithm can be divided into two separate parts:
\begin{enumerate}
    \item Simulation --- The simulation part is responsible for generating the time trajectories of the system. In each time step, the algorithm decides the next trajectory step by evaluating each node's boolean formula. The results of the formulas are then used to update the node state.
    \item Statistics computation --- The statistics computation takes raw trajectory data from the simulation part and transforms it to be more visualization-friendly. It envelopes the probability distribution of trajectory states across multiple trajectory samples and the computation of the final and fixed states of the system.
\end{enumerate}

\subsubsection*{Simulation}

The simulation part is the most computationally demanding part of the CPU MaBoSS implementation. The preliminary performance reports show that around $30-80\%$ of the MaBoSS runtime (depending on the model) is spent in the evaluation of Boolean formulas. This evaluation is carried out in a recursive call over an expression tree. This approach is not well suited for a CPU cache because the whole tree may span multiple cache lines; its traversal may easily result in a major cache miss ratio. The GPUs are even more susceptible to this problem because they have a lower cache per core ratio. We took this as a starting point for our optimizations.

There are multiple ways to tackle the problem of bad cache performance of the expression tree. The first one is to use a different data structure that is more cache-friendly, such as the van Emde Boas tree layout. In our approach, we take one step further by utilizing the \emph{runtime compilation of GPU code}. Using this technique, instead of building the expression tree for the Boolean formulas, we compile the formulas into a native binary code that can be executed directly on the GPU. The main advantage of this approach is that the formulas are encoded in the instructions instead of being stored in the memory. This way, the formulas would not need to be fetched from memory multiple times during the application runtime. The other advantage is that the compiler can optimize the GPU code to utilize the GPU hardware more efficiently.

The major drawback of the runtime compilation is its speed. The compilation takes magnitudes longer than the building of the expression trees. On the other hand, the formulas compilation is executed only once at the beginning of the simulation and the compiled code is reused for the whole simulation. For the big enough models, the compilation time gets amortized well by the overall simulation and statistics computation time.

\subsubsection*{Statistics computation}

The CPU implementation of this part of the algorithm heavily relies on the usage of hash maps. They are used to store node states as the keys, and the values may be the state occurrences or their average timespan in the trajectories, depending on the specific statistics computation. We will focus on the probability distribution of trajectory states since it is the most computationally demanding part of the statistics computation and the optimizations used for this part apply to the other statistics computations as well.

The optimization of this step relies on the fact that the typical real-life MaBoSS models have many \emph{internal nodes}. A node is marked internal if we do not want it to contribute to the statistics computation. As a result, the typical number of nodes that are not internal is much smaller than the total number of nodes; e.g., in the model with 1000 nodes and only 5 non-internal nodes, the number of possible states in a trajectory can be $2^{1000}$, but the number of states that are important for the statistics computation is only $2^5$.

We used this fact to substitute the hash table with a more GPU-friendly histogram. The relatively low number of states generated by non-internal nodes allows us to materialize the whole histogram as a fixed-size array. The states are no longer stored as the keys --- rather, a hash function is used to map the states to the histogram array indices. We use a simple hash function with bit masking and shifting to extract and compact non-internal nodes from the state representation. Further, we use well-known GPU histogram optimizations, such as shared memory privatization and atomic operations, to improve the performance of the histogram computation.

\section*{Results}

\subsection*{MaBoSS.MPI}
We developed an MPI parallel implementation of MaBoSS.

We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}). A comparison of execution times of MaBoSS v2.5.3 and v2.4.0 using 40 cores and one million individual simulations showed a considerable (3×) speed-up between the pre- and post-optimisation versions, which is mostly due to optimised memory usage (figure 2, purple vs red). We used the same test with the MPI implementation of MaBoSS, using 20 cores for each MPI node, from 1 MPI node up to 192 MPI nodes. We can observe a very good speedup (>50x) up to 64 MPI nodes (1.28k cores), and a plateau for larger setups (Figure 7, green). This is probably due the very short simulation time (10s) for that setup and the minimum overhead for such a parallel simulation, and needs testing on larger simulations.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{Figures/sizek_model.pdf}
\caption{Scalability results for Sizek model}
\label{fig:sizek_results}
\end{figure}

In order to explore larger model sizes, we created a synthetic model with 1000 nodes (Figure~\ref{fig:synthetic_results}). We simulated this model on 32 cores per MPI node, on 1 to 192 MPI nodes (32 to 6144 cores). With this dataset, the simulation time goes from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As we hypothesised, with larger simulation times, a decline in the speedup was not observed as in the previous example 
The MaBoSS benchmarking measurements described in this section were performed on MareNostrum 4.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
\caption{Scalability results for synthetic model with 1000 nodes}
\label{fig:synthetic_results}
\end{figure}

\subsection*{MaBoSS.GPU}
MaBoSSG is written in C++/CUDA. It executes the MaBoSS time trajectories simulation on GPUs outputting millions of simulated trajectories per second.

MaBoSSG implements subset of MaBoSS algorithms over the time trajectories: The probabilities estimates and final+fixed states discovery.

%\begin{figure}%[tbhp]
%\centering
%\includegraphics[width=.8\linewidth]{Figures/Figure_1}
%\caption{Placeholder image of Iris with a long example caption to show justification setting.}
%\label{fig:computerNo}
%\end{figure}

\section*{Conclusions}

We did great work\\

\subsection*{Future works} 
bLA, BLA BLA

\begin{acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
\end{acknowledgements}

\begin{funding}
This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
\end{funding}

\section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
\onecolumn
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supplementary Information %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\captionsetup*{format=largeformat}
%\section{Something about something} \label{note:Note1} 
%Something

%TC:endignore
%the command above ignores this section for word count

\end{document}

### 2023-12-17 Article.tex
update implementation according to the new sections

@@
 For each node, a \emph{Boolean logic} $\mathcal{B}: \mathcal{S} \rightarrow [0, \infty)$ is specified.
-The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$ (where only $S_i$ and $S'_i$ differ).
+The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$, where only $S_i$ and $S'_i$ differ.
 If $\mathcal{B}_i(S) = 0$, then the transition is not allowed.
@@
 Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective.
-A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and binary expression trees, that are not well suited for GPU hardware.
-In this section, we describe a CUDA GPU implementation of MaBoSS, and we detail the optimizations used to enable higher performance.
-
-The MaBoSS algorithm can be divided into two separate parts:
-\begin{enumerate}
-    \item Simulation --- The simulation part is responsible for generating the time trajectories of the system.
-In each time step, the algorithm decides the next trajectory step by evaluating each node's boolean formula.
-The results of the formulas are then used to update the node state.
-    \item Statistics computation --- The statistics computation takes raw trajectory data from the simulation part and transforms it to be more visualization-friendly.
-It envelopes the probability distribution of trajectory states across multiple trajectory samples and the computation of the final and fixed states of the system.
-\end{enumerate}
+A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware.
+In this section, we describe a CUDA GPU implementation of the simulation step and the stats computation steps, and we detail the optimizations used to enable higher performance.
 
@@
 
+In our implementation, we decided to assign a single GPU thread to a single trajectory.
+We mentioned in the previous section, that a work distribution, where multiple threads would work on a single trajectory, may be beneficial for large models.
+However, this can not be implemented trivially because this would introduce a thread divergence in the transition rates computation, which is penalized by a serial execution of the GPU threads.
+This optimization requires a clever rewrite of the Boolean logic, which we leave for future work.
+
 \subsubsection*{Statistics computation}
@@
 
-The optimization of this step relies on the fact that the typical real-life MaBoSS models have many \emph{internal nodes}.
-A node is marked internal if we do not want it to contribute to the statistics computation.
-As a result, the typical number of nodes that are not internal is much smaller than the total number of nodes; e.g., in the model with 1000 nodes and only 5 non-internal nodes, the number of possible states in a trajectory can be $2^{1000}$, but the number of states that are important for the statistics computation is only $2^5$.
-
-We used this fact to substitute the hash table with a more GPU-friendly histogram.
+The optimization of this step relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes regardless of the total number of nodes.
 The relatively low number of states generated by non-internal nodes allows us to materialize the whole histogram as a fixed-size array.

### 2023-12-21 Article.tex
add results section

@@
 
-\subsection*{MaBoSS.GPU}
-MaBoSSG is written in C++/CUDA. It executes the MaBoSS time trajectories simulation on GPUs outputting millions of simulated trajectories per second.
-
-MaBoSSG implements subset of MaBoSS algorithms over the time trajectories: The probabilities estimates and final+fixed states discovery.
-
 %\begin{figure}%[tbhp]
@@
 %\end{figure}
+
+\subsection*{MaBoSS.GPU}
+
+In this section, we present the performance results of MaBoSS.GPU by comparing the runtimes of the CPU and GPU implementations on real-world and synthetic models.
+We also take a deeper look into the implementation by analyzing the overhead of the runtime compilation of the Boolean formulas.
+
+\subsubsection*{Benchmarking Methodology}
+
+For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$.
+In order to test the scalability of the GPU implementation, we also created a synthetic models with up to $1000$ nodes.
+Synthetic models are designed in a way such that the length of each trajectory is predictable, and the models have no stable states.
+Also, the number of non-internal nodes is kept low (10 nodes) to enable the usage of the histogram optimization.
+
+The GPU implementation benchmarks were run on a datacenter-grade NVIDIA Tesla A100 GPU and a consumer-grade NVIDIA RTX 3070 Laptop GPU.
+The CPU implementation benchmarks were run on a 32-core Intel Xeon Gold 6130 CPU with multithreading.
+The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2.
+Each measurement was repeated $10$ times, and the average runtime was used for the final results after removing the outliers.
+
+\subsubsection*{Measurements}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{Figures/real.pdf}
+    \caption{Wall time comparison of CPU and GPU implementations on real-world models. Each model is simulated with 1 million trajectories.}
+    \label{fig:real}
+\end{figure}
+
+In Figure~\ref{fig:real}, we compare the wall time of the CPU and GPU implementations on real-world datasets.
+The GPU implementation is faster than the CPU implementation on all models, and the speedup shows to be more significant on the models with more nodes and longer trajectories.
+On Montagud model with 133 nodes but a relatively short average trajectory, we achieve $145\times$ speedup.
+On a slightly smaller Sizek model with a longer average trajectory, the speedup is up to $326\times$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{Figures/nodes.pdf}
+    \caption{Wall time comparison of CPU and GPU implementations on synthetic models. Each model is simulated with 1 million trajectories.}
+    \label{fig:synth}
+\end{figure}
+
+Figure~\ref{fig:synth} shows much finer performance progression on synthetic models.
+On the left subplot, the runtime compilation is filtered out to show the pure simulation performance.
+We can see that the CPU variant starts to progress steeper at around the $100$ nodes boundary.
+We can expect that the implementation hits the cache size limit and fetching required data from the memory comes with more overhead.
+The same can be observed in the GPU variant, but at $200$ nodes and with a much more drastic increase.
+This observation corresponds to a much more significant performance penalty when fetching data from the GPU memory.
+Overall, this suggests that the optimization of dividing transition rates computation among multiple threads mentioned in the previous section may provide a significant speedup for bigger models as it alleviates the register pressure.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{Figures/nodes-compilation-big-NVIDIA Tesla A100 GPU.pdf}
+    \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
+    \label{fig:comp}
+\end{figure}
+
+The right subplot shows the total runtime of the GPU implementation.
+Comparing the subplots shows that the runtime compilation ratio decreases with increasing model size.
+Figure~\ref{fig:comp} offers more detailed benchmarks for this scenario, run on the NVIDIA Tesla A100 GPU.
+The compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).
+Notably, as soon as the simulation becomes more complex, i.e., by increasing the number of nodes or simulated trajectories, the compilation time becomes less than $10\%$ of the total runtime, even for the models with unrealistically long formulae.
+This suggests that the runtime compilation is a viable optimization for bigger models.
 

### 2024-02-11 Article.tex
overleaf sync

@@
 \author[2]{Miroslav Kratochvíl}
+\author[3,4,5]{Emmanuel Barillot}
+\author[3,4,5]{Laurence Calzone}
 \author[3,4,5,\Letter]{Vincent Noël}
@@
 \section*{Introduction}
-
-MaBoSS \cite{stoll2012continuous, stoll2017maboss}\\
-Why Boolean models are important\\
-Why are they getting larger\\
-Why are they need to be adapted for HPC environment\\
+Biological systems are very large and complex, and understanding their behavior remains a very important task to allow the design of new therapies for complex diseases such as cancer.
+A key approach in this endeavor is to build computational models from existing knowledge, and to analyze them to try to find intervention points and predict the efficacy of new treatment.
+Many different frameworks have been used to describe biological systems, from very quantitative systems of differential equations, to more qualitative approaches such as Boolean models.
+While the former seems more adapted to represent complex behavior such as non-linear dependencies, the latter is increasingly used because it allows the representation of very large systems.
+Many Boolean models have been built to describe biological systems, to tackle a variety of problems : from understanding basic properties of cell cycle\citep{faure2006cellcycle,sizek2019boolean}, to more advanced properties of cancer \citep{montagud2022prostate} or ...
+Historically, the task of building such models involved reading a very large literature and summarizing it in a list of important components and their interactions.
+More recently, thanks to advances in databases listing such interactions\citep{licata2020signor,turei2016omnipath}, and to experimental techniques providing informations on a larger number of components [rnaseq, single cell, …], methods have been designed to automatically infer Boolean formulas from the constraints encoded in the knowledge and in the experimental data \cite{chevalier2020synthesis,benevs2023boolean} to build large Boolean models.
+While this effort suffers many challenges, we believe it is the way to go to study the large scale complexity of biological systems.
+In order to analyse the dynamical properties of such large Boolean models, we need efficient methods, among them for simulations.
+In this effort, adapting to the modern HPC computing architectures existing softwares in an important goal.
+MaBoSS \cite{stoll2012continuous, stoll2017maboss} is a stochastic Boolean simulator which performs estimations of state probability trajectories, based on Markov chains. 
+In this paper, we discuss about the adaptation of MaBoSS simulation algorithm for two different hardware architectures : large CPU clusters, with a MPI parallel implementation, and GPU accelerators.
 
@@
 For each node, a \emph{Boolean logic} $\mathcal{B}: \mathcal{S} \rightarrow [0, \infty)$ is specified.
-The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$, where only $S_i$ and $S'_i$ differ.
+The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$ (where only $S_i$ and $S'_i$ differ).
 If $\mathcal{B}_i(S) = 0$, then the transition is not allowed.
@@
 \subsection*{MaBoSS.MPI}
-
+MaBoSS' CPU implementation already offered parallelism, allowing to use multiple CPU cores to compute the multiple trajectories.
+However, this implementation stays limited to a single machine CPU, and its scalability suffers from a very intensive memory usage.
+Furthermore, with the increasing number of nodes in Boolean models, the need for very large number of trajectories leads to large simulation time.
+To tackle this, we decided to extend this implementation and add another layer of parallelism using MPI.
+In this implementation, the simulation of the trajectories is not only distributed on multiple cores, but on the multiple cores of multiple MPI nodes.
+By using this implementation, we are able to exploit the large CPU clusters available in HPC centers.
+The implementation of this MPI implementation is straightforward, and consists in distributing the invidivual trajectories even further, first on MPI nodes and then on the CPU cores of the MPI node.
+Once the trajectories are simulated, the statistics are first computed for each MPI nodes, and then the global results is computer from the results of each MPI nodes.
 \subsection*{MaBoSS.GPU}
@@
 \subsection*{MaBoSS.MPI}
-We developed an MPI parallel implementation of MaBoSS.
-
 We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}).
@@
 \begin{acknowledgements}
-We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+We thank Gautier Stoll for his guidance and fruitful discussion.
 \end{acknowledgements}

### 2024-02-11 Article.tex
rewrite intro; add outro; better connect MPI

--- 2023-12-21 Article.tex
@@
 \author[2]{Miroslav Kratochvíl}
+\author[3,4,5]{Emmanuel Barillot}
+\author[3,4,5]{Laurence Calzone}
 \author[3,4,5,\Letter]{Vincent Noël}
 
-\affil[1]{Department of Software Engineering, Charles University, Prague, Czech Republic}
+\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
 \affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
@@
 \section*{Introduction}
+Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer.
+A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
+Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models.
+While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
+Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{montagud2022prostate}.
 
-MaBoSS \cite{stoll2012continuous, stoll2017maboss}\\
-Why Boolean models are important\\
-Why are they getting larger\\
-Why are they need to be adapted for HPC environment\\
+Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
+More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and experimental techniques providing information on a bigger number of components [rnaseq, single cell, …], the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{chevalier2020synthesis,benevs2023boolean} to build large Boolean models.
+While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems.
+Furthermore, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation methods.
+In this effort, adapting existing software to modern HPC computing architectures is an important goal.
+
+Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
+This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation.
 
@@
 
-In this section, we describe the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.
+We begin with describing the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.
 
@@
 This puts a big pressure on the choice of the data representation for the state $S$ and Boolean logic formulas.
-We discuss this topic in the next section.
+We discuss this topic in the Implementation section.
 
@@
 
-In the following section, we describe the implementation of MaBoSS.MPI and MaBoSS.GPU. We use this discussion to detail the specific optimization choices in the implementation of the algorithms.
+Finally, we use this discussion to detail the specific optimization choices in the implementation of the algorithms.
 
-\section*{Implementation}
-\subsection*{MaBoSS.MPI}
+\section*{Implementation}\label{sec:implementation}
+
+In this section, we describe the high-level overview of MaBoSS.GPU and MaBoSS.MPI implementation.
 
@@
 A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware.
-In this section, we describe a CUDA GPU implementation of the simulation step and the stats computation steps, and we detail the optimizations used to enable higher performance.
+We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.
 
@@
 
+\subsection*{MaBoSS.MPI}
+Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code.
+Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory.
+The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.
+
+In summary, MaBoSS.MPI evenly distributes individual trajectories first on MPI nodes and then on the corresponding CPU cores, assuming the MPI nodes are homogenous.
+Once the trajectories are simulated and the statistics are computed for each MPI node, the intermediate data are reduced into a final result using MPI collective operations.
+
+% MaBoSS' CPU implementation already offered parallelism, allowing to use multiple CPU cores to compute the multiple trajectories.
+However, this implementation stays limited to a single machine CPU, and its scalability suffers from a very intensive memory usage.
+Furthermore, with the increasing number of nodes in Boolean models, the need for very large number of trajectories leads to large simulation time.
+% To tackle this, we decided to extend this implementation and add another layer of parallelism using MPI.
+In this implementation, the simulation of the trajectories is not only distributed on multiple cores, but on the multiple cores of multiple MPI nodes.
+By using this implementation, we are able to exploit the large CPU clusters available in HPC centers.
+% The implementation of this MPI implementation is straightforward, and consists in distributing the invidivual trajectories even further, first on MPI nodes and then on the CPU cores of the MPI node.
+Once the trajectories are simulated, the statistics are first computed for each MPI nodes, and then the global results is computer from the results of each MPI nodes.
+
 \section*{Results}
 
-\subsection*{MaBoSS.MPI}
-We developed an MPI parallel implementation of MaBoSS.
-
-We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}).
-A comparison of execution times of MaBoSS v2.5.3 and v2.4.0 using 40 cores and one million individual simulations showed a considerable (3×) speed-up between the pre- and post-optimisation versions, which is mostly due to optimised memory usage (figure 2, purple vs red).
-We used the same test with the MPI implementation of MaBoSS, using 20 cores for each MPI node, from 1 MPI node up to 192 MPI nodes.
-We can observe a very good speedup (>50x) up to 64 MPI nodes (1.28k cores), and a plateau for larger setups (Figure 7, green).
-This is probably due the very short simulation time (10s) for that setup and the minimum overhead for such a parallel simulation, and needs testing on larger simulations.
-
-\begin{figure}%[tbhp]
-\centering
-\includegraphics[width=.8\linewidth]{Figures/sizek_model.pdf}
-\caption{Scalability results for Sizek model}
-\label{fig:sizek_results}
-\end{figure}
-
-In order to explore larger model sizes, we created a synthetic model with 1000 nodes (Figure~\ref{fig:synthetic_results}).
-We simulated this model on 32 cores per MPI node, on 1 to 192 MPI nodes (32 to 6144 cores).
-With this dataset, the simulation time goes from 20 hours on 1 MPI node to 430 seconds on 192 nodes.
-As we hypothesised, with larger simulation times, a decline in the speedup was not observed as in the previous example
-The MaBoSS benchmarking measurements described in this section were performed on MareNostrum 4.
-
-\begin{figure}%[tbhp]
-\centering
-\includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
-\caption{Scalability results for synthetic model with 1000 nodes}
-\label{fig:synthetic_results}
-\end{figure}
-
-%\begin{figure}%[tbhp]
-%\centering
-%\includegraphics[width=.8\linewidth]{Figures/Figure_1}
-%\caption{Placeholder image of Iris with a long example caption to show justification setting.}
-%\label{fig:computerNo}
-%\end{figure}
-
-\subsection*{MaBoSS.GPU}
-
-In this section, we present the performance results of MaBoSS.GPU by comparing the runtimes of the CPU and GPU implementations on real-world and synthetic models.
-We also take a deeper look into the implementation by analyzing the overhead of the runtime compilation of the Boolean formulas.
+To evaluate the discussed optimizations, we present the results of performance benchmarks for MaBoSS.GPU and MaBoSS.MPI by comparing their runtimes against the original CPU implementation.
+To get the most accurate results, we used real-world models and synthetic models with varying sizes.
 
@@
 For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$.
-In order to test the scalability of the GPU implementation, we also created a synthetic models with up to $1000$ nodes.
+In order to test the scalability of the GPU and MPI implementation, we also created a synthetic models with up to $1000$ nodes.
 Synthetic models are designed in a way such that the length of each trajectory is predictable, and the models have no stable states.
@@
 
-\subsubsection*{Measurements}
+The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer.
+
+\subsection*{MaBoSS.GPU}
 
@@
 
+\subsection*{MaBoSS.MPI}
+
+\begin{figure}%[tbhp]
+\centering
+\includegraphics[width=.8\linewidth]{Figures/sizek_model.pdf}
+\caption{Scalability results for Sizek model}
+\label{fig:sizek_results}
+\end{figure}
+
+Figure~\ref{fig:sizek_results} shows the efficiency of the MPI implementation on the Sizek model.
+We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores.
+We can observe a close-to-linear speedup of up to 64 MPI nodes (1.28k cores), and a plateau for larger suites (Figure 7, green).
+This can be explained by hitting a bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.
+
+\begin{figure}%[tbhp]
+\centering
+\includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
+\caption{Scalability results for synthetic model with 1000 nodes}
+\label{fig:synthetic_results}
+\end{figure}
+
+To stress the scalability of the implementation, we also used a synthetic model with 1000 nodes (see Figure~\ref{fig:synthetic_results}).
+We simulated this model on 32 cores per MPI node, on 1 to 192 nodes (32 to 6144 cores).
+With this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes.
+As we hypothesized, a decline in the speedup was not observed in larger simulations.
+
+% We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}).
+A comparison of execution times of MaBoSS v2.5.3 and v2.4.0 using 40 cores and one million individual simulations showed a considerable (3×) speed-up between the pre- and post-optimisation versions, which is mostly due to optimised memory usage (figure 2, purple vs red).
+We used the same test with the MPI implementation of MaBoSS, using 20 cores for each MPI node, from 1 MPI node up to 192 MPI nodes.
+We can observe a very good speedup (>50x) up to 64 MPI nodes (1.28k cores), and a plateau for larger setups (Figure 7, green).
+This is probably due the very short simulation time (10s) for that setup and the minimum overhead for such a parallel simulation, and needs testing on larger simulations.
+
+% In order to explore larger model sizes, we created a synthetic model with 1000 nodes (Figure~\ref{fig:synthetic_results}).
+We simulated this model on 32 cores per MPI node, on 1 to 192 MPI nodes (32 to 6144 cores).
+With this dataset, the simulation time goes from 20 hours on 1 MPI node to 430 seconds on 192 nodes.
+As we hypothesised, with larger simulation times, a decline in the speedup was not observed as in the previous example
+% The MaBoSS benchmarking measurements described in this section were performed on MareNostrum 4.
+
+% \begin{figure}%[tbhp]
+% \centering
+% \includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
+% \caption{Scalability results for synthetic model with 1000 nodes}
+% \label{fig:synthetic_results}
+% \end{figure}
+
+%\begin{figure}%[tbhp]
+%\centering
+%\includegraphics[width=.8\linewidth]{Figures/Figure_1}
+%\caption{Placeholder image of Iris with a long example caption to show justification setting.}
+%\label{fig:computerNo}
+%\end{figure}
+
 \section*{Conclusions}
 
-We did great work\\
+In this work, we presented two new implementations of MaBoSS, a continuous time Boolean model simulator, for HPC environments: the first implementation, MaBoSS.GPU, is designed to exploit the computational power of massively parallel GPU hardware, and the second implementation, MaBoSS.MPI, which enables MaBoSS to scale to many nodes of HPC clusters.
+We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both implementations are capable of providing significant speedups over the original CPU implementation.
+The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.
 
-\subsection*{Future works} 
-bLA, BLA BLA
+Overall, we believe the new MaBoSS extensions will enable the simulation of large, automatically generated models and will be a valuable tool for the systems biology community.
+
+\subsection*{Future work}
+
+As mentioned in the discussion section, the next steps in the development of MaBoSS.GPU would be to parallelize the iteration step of the simulation.
+This requires a different representation of Boolean formulas, which would not pose any thread divergence, probably by substituting runtime compilation with an in-memory data structure.
+This optimization could potentially decrease the register pressure created by holding the state data and thus increase the performance of huge models with hundreds of nodes.
+
+Furthermore, many parallelization methods used in GPU implementation could be backported to the CPU implementation, increasing the overall usability of MaBoSS running on systems without GPU accelerators.
+
+Finally, to enable the simulation of extra-large models with thousands of nodes, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.
 
 \begin{acknowledgements}
-We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+We thank Gautier Stoll for his guidance and fruitful discussion.
 \end{acknowledgements}

### 2024-02-11 Article.tex
add some refs

@@
 There are multiple ways to tackle the problem of bad cache performance of the expression tree.
-The first one is to use a different data structure that is more cache-friendly, such as the van Emde Boas tree layout.
+The first one is to use a different data structure that is more cache-friendly, such as the van Emde Boas tree layout~\cite{van1975preserving}.
 In our approach, we take one step further by utilizing the \emph{runtime compilation of GPU code}.
@@
 
-For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$.
+For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$ (\texttt{cellcycle}~\cite{faure2006cellcycle}, \texttt{sizek}~\cite{sizek2019boolean} and \texttt{Montagud}~\cite{montagud2022prostate}).
 In order to test the scalability of the GPU and MPI implementation, we also created a synthetic models with up to $1000$ nodes.

### 2024-02-14 Article.tex
slight revision

@@
 For each node, a \emph{Boolean logic} $\mathcal{B}: \mathcal{S} \rightarrow [0, \infty)$ is specified.
-The logic of $i$-th node determines if a state $S$ is allowed to transition to a state $S'$ (where only $S_i$ and $S'_i$ differ).
+The logic of $i$-th node determines if a state $S$ is allowed to transition to such state $S'$, where only $S_i$ and $S'_i$ differ.
 If $\mathcal{B}_i(S) = 0$, then the transition is not allowed.
@@
 Let us start by analyzing the time complexity of one iteration step of the simulation.
-Let us have a model with $n$ nodes.
+Suppose a model with $n$ nodes.
 If we assume that one Boolean logic formula can be evaluated in $\mathcal{O}(n)$, then the computation of transition rates can be done in $\mathcal{O}(n^2)$.
@@
 There is a clear dependency between the trajectory iterations, so $\mathcal{O}(u)$ steps must be performed serially.
-Realistically, parallelizing an iteration step may prove beneficial only for rather large models such that the scheduling overhead and communications cost are lower than the achieved parallelism.
+Realistically, parallelizing an iteration step may prove beneficial only for rather large models when the scheduling overhead and communications cost are lower than the achieved parallelism.
 
@@
 This evaluation is carried out in a recursive call over an expression tree.
-This approach is not well suited for a CPU cache because the whole tree may span multiple cache lines; its traversal may easily result in a major cache miss ratio.
+This approach is not well suited for a CPU cache because the whole tree may span multiple cache lines and, apart from other issues, its traversal may easily result in a major cache miss ratio~\cite{karlsson2000prefetching}.
 The GPUs are even more susceptible to this problem because they have a lower cache per core ratio.
@@
 There are multiple ways to tackle the problem of bad cache performance of the expression tree.
-The first one is to use a different data structure that is more cache-friendly, such as the van Emde Boas tree layout~\cite{van1975preserving}.
-In our approach, we take one step further by utilizing the \emph{runtime compilation of GPU code}.
+One can either use a linked data structure that is more cache-friendly, such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array with an ad-hoc layout.
+In our approach, we take one step further by utilizing the \emph{runtime compilation of GPU code}~\cite{nvrtc}.
 Using this technique, instead of building the expression tree for the Boolean formulas, we compile the formulas into a native binary code that can be executed directly on the GPU.
@@
 On the other hand, the formulas compilation is executed only once at the beginning of the simulation and the compiled code is reused for the whole simulation.
-For the big enough models, the compilation time gets amortized well by the overall simulation and statistics computation time.
+In theory, the compilation time can get amortized well by the overall simulation and statistics computation time for big enough models.
 
@@
 We mentioned in the previous section, that a work distribution, where multiple threads would work on a single trajectory, may be beneficial for large models.
-However, this can not be implemented trivially because this would introduce a thread divergence in the transition rates computation, which is penalized by a serial execution of the GPU threads.
+However, this can not be implemented trivially without creating a thread divergence in the transition rates computation (which is usually penalized by a completely serialized execution anyway).
 This optimization requires a clever rewrite of the Boolean logic, which we leave for future work.
@@
 To evaluate the discussed optimizations, we present the results of performance benchmarks for MaBoSS.GPU and MaBoSS.MPI by comparing their runtimes against the original CPU implementation.
-To get the most accurate results, we used real-world models and synthetic models with varying sizes.
+To get comprehensive results, we used real-world models and synthetic models with varying sizes.
 
@@
 We can see that the CPU variant starts to progress steeper at around the $100$ nodes boundary.
-We can expect that the implementation hits the cache size limit and fetching required data from the memory comes with more overhead.
-The same can be observed in the GPU variant, but at $200$ nodes and with a much more drastic increase.
+We can expect that the implementation hits the cache size limit and fetching the required data from the memory comes with more overhead.
+The same can be observed in the GPU variant, but at $200$ nodes and with more drastic increase.
 This observation corresponds to a much more significant performance penalty when fetching data from the GPU memory.
@@
 We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores.
-We can observe a close-to-linear speedup of up to 64 MPI nodes (1.28k cores), and a plateau for larger suites (Figure 7, green).
+We can observe a close-to-linear speedup of up to 64 MPI nodes (1280 cores), and a plateau for larger suites (Figure~\ref{fig:sizek_results}, green).
 This can be explained by hitting a bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.

### 2024-02-14 Article.tex
Improved introduction, added repos urls

@@
 While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
-Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{montagud2022prostate}.
+Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.
 
 Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
-More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and experimental techniques providing information on a bigger number of components [rnaseq, single cell, …], the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{chevalier2020synthesis,benevs2023boolean} to build large Boolean models.
+More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean} to build large Boolean models.
 While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems.
-Furthermore, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation methods.
+However, in order to analyze the dynamic properties of such large Boolean models, we need to efficient simulation tools.
 In this effort, adapting existing software to modern HPC computing architectures is an important goal.
@@
 This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation.
+The source codes are available at \url{https://github.com/sysbio-curie/MaBoSSG/} and \url{https://github.com/sysbio-curie/MaBoSS-env-2.0/}, together with associated benchmarking data.
 

### 2024-02-14 Article.tex
add mpi plots

@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/sizek_model.pdf}
-\caption{Scalability results for Sizek model}
+\includegraphics[width=.8\linewidth]{Figures/sizek_mpi.pdf}
+\caption{Scalability results of MPI implementation on Sizek model with 20 cores per MPI node.}
 \label{fig:sizek_results}
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
-\caption{Scalability results for synthetic model with 1000 nodes}
+\includegraphics[width=.8\linewidth]{Figures/synth_mpi_speedup.pdf}
+\caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and 32 cores per MPI node.}
 \label{fig:synthetic_results}

### 2024-02-14 mpi-plots/plots.R

@@ -1,0 +1,214 @@
+
+if(F) {
+  # use this to install deps
+  install.packages(c('ggplot2', 'cowplot', 'sitools'))
+}
+
+library(ggplot2)
+library(cowplot)
+
+# x <- read.table('data.tsv', header=T)
+
+# #
+# # helpers
+# #
+
+# sival <- function(...) Vectorize(function(x)sitools::f2si(x, ...))
+
+# labeldf <- function(df) {
+#   for(x in names(df))
+#     df[[x]] <- paste(x, '=', df[[x]])
+#   df
+# }
+
+# #
+# # Scaling: how much resources do we need to solve a problem of given expected
+# # size? (is there an unexpected scalability problem?)
+# #
+
+# ggsave("plots/scalability.png", units="in", width=7, height=7,
+# ggplot(x, aes(size, time*cpus/size, color=algorithm)) +
+#   geom_point() +
+#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
+#   scale_x_log10("Base problem size", labels=sival('op')) +
+#   scale_y_log10("Time per operation", labels=sival('Cs/op')) +
+#   ggtitle("Overhead of processing different problem sizes") +
+#   scale_color_brewer("Software", palette='Dark2') +
+#   scale_fill_brewer("Software", palette='Dark2') +
+#   theme_cowplot(font_size=9) +
+#   theme(
+#     panel.grid.major=element_line(size=.2, color='#cccccc'),
+#   )
+# )
+
+# #
+# # Speed: How much does throwing extra CPUs help to solve the problem faster?
+# #
+
+# ggsave("plots/speed.png", units="in", width=7, height=7,
+# ggplot(x, aes(cpus/size, time/size, color=algorithm)) +
+#   geom_point() +
+#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
+#   scale_x_log10("Base operations assigned to one CPU", labels=function(x)(sival('op/C')(1/x))) +
+#   scale_y_log10("Time per base operation", labels=sival('s/op')) +
+#   ggtitle("Speedup given for invested resources") +
+#   scale_color_brewer("Software", palette='Dark2') +
+#   scale_fill_brewer("Software", palette='Dark2') +
+#   theme_cowplot(font_size=9) +
+#   theme(
+#     panel.grid.major=element_line(size=.2, color='#cccccc'),
[…]

### 2024-02-14 Article.tex
giga format

@@
-\RequirePackage[2020-02-02]{latexrelease}
-\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
+\documentclass[a4paper,num-refs]{oup-contemporary}
+% \RequirePackage[2020-02-02]{latexrelease}
+% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
+
+\journal{gigascience}
+
+\usepackage{graphicx}
+\usepackage{siunitx}
+
 %\usepackage{blindtext}
@@
 
-% Please give the surname of the lead author for the running footer
-\leadauthor{Šmelko}
-
-\begin{document}
-
 \title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
-\shorttitle{MaBoSS for HPC environments}
+% \shorttitle{MaBoSS for HPC environments}
 
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[1,\Letter]{Adam Šmelko}
+\author[1,\authfn{1}]{Adam Šmelko}
 \author[2]{Miroslav Kratochvíl}
 \author[3,4,5]{Emmanuel Barillot}
-\author[3,4,5]{Laurence Calzone}
-\author[3,4,5,\Letter]{Vincent Noël}
+% \author[3,4,5]{Laurence Calzone}
+\author[3,4,5,\authfn{1}]{Vincent Noël}
 
@@
 \affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
+
+\authnote{\authfn{1}smelko@d3s.mff.cuni.cz; vincent.noel@curie.fr}
+
+\papercat{Paper}
+
+\runningauthor{Šmelko et al.}
+
+\jvolume{00}
+\jnumber{0}
+\jyear{2023}
+
+\begin{document}
+
+\begin{frontmatter}
 \maketitle
@@
 \end{keywords}
+\end{frontmatter}
 
-\begin{corrauthor}
-smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
-\end{corrauthor}
-\section*{Introduction}
+\section{Introduction}
 Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer.
@@
 While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems.
-However, in order to analyze the dynamic properties of such large Boolean models, we need to efficient simulation tools.
+However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation tools.
 In this effort, adapting existing software to modern HPC computing architectures is an important goal.
 
-Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
-This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation.
-The source codes are available at \url{https://github.com/sysbio-curie/MaBoSSG/} and \url{https://github.com/sysbio-curie/MaBoSS-env-2.0/}, together with associated benchmarking data.
+Our work focuses on extending MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
+The main contributions of this paper are twofold:
+\begin{itemize}
+    \item MaBoSS.GPU, a GPU-accelerated implementation of MaBoSS, which is designed to exploit the computational power of massively parallel GPU hardware.
+    \item MaBoSS.MPI, a parallel implementation of MaBoSS which can scale to multinode environments, such as large CPU clusters.
+\end{itemize}
 
-\section*{Background}
+The source codes of the proposed implementations are publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}.
+We also provide the scripts, presented graphs, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/asmelko/gigascience24-artifact}}.
+
+The paper is organized as follows.
+In the next section, we discuss the MaBoSS algorithm and its complexity.
+In the following section, we detail the implementation of MaBoSS.GPU and MaBoSS.MPI.
+In the results section, we present the performance benchmarks of the new implementations.
+Finally, we conclude the paper and discuss the future work.
+[TODO - add section refs]
+
+\section{Background}
 
@@
 
-\section*{Complexity evaluation and parallelization opportunities of MaBoSS}
+\section{Complexity evaluation and parallelization opportunities of MaBoSS}
 
@@
 
-\subsection*{Simulation complexity and parallelization}
+\subsection{Simulation complexity and parallelization}
 
@@
 
-\subsection*{Statistics computation complexity and parallelization}
+\subsection{Statistics computation complexity and parallelization}
 
@@
 
-\section*{Implementation}\label{sec:implementation}
+\section{Implementation}\label{sec:implementation}
 
@@
 
-\subsection*{MaBoSS.GPU}
+\subsection{MaBoSS.GPU}
 Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective.
@@
 
-\subsubsection*{Simulation}
+\subsubsection{Simulation}
 
@@
 
-\subsubsection*{Statistics computation}
+\subsubsection{Statistics computation}
 
@@
 
-\subsection*{MaBoSS.MPI}
+\subsection{MaBoSS.MPI}
 Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code.
@@
 
-\section*{Results}
+\section{Results}
 
@@
 
-\subsubsection*{Benchmarking Methodology}
+\subsection{Benchmarking Methodology}
 
@@
 
-\subsection*{MaBoSS.GPU}
+\subsection{MaBoSS.GPU}
 
@@
 
-\subsection*{MaBoSS.MPI}
+\subsection{MaBoSS.MPI}
 
@@
 
-\section*{Conclusions}
+\section{Conclusions}
 
-In this work, we presented two new implementations of MaBoSS, a continuous time Boolean model simulator, for HPC environments: the first implementation, MaBoSS.GPU, is designed to exploit the computational power of massively parallel GPU hardware, and the second implementation, MaBoSS.MPI, which enables MaBoSS to scale to many nodes of HPC clusters.
-We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both implementations are capable of providing significant speedups over the original CPU implementation.
+In this work, we presented two new implementations of MaBoSS tool, a continuous time Boolean model simulator, designed for HPC environments:  MaBoSS.GPU which is designed to exploit the computational power of massively parallel GPU hardware, and MaBoSS.MPI which enables MaBoSS to scale to many nodes of HPC clusters.
+We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both variants are capable of providing significant speedups over the original CPU code.
 The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.
@@
 
-\subsection*{Future work}
+\subsection{Future work}
 
@@
 
-Finally, to enable the simulation of extra-large models with thousands of nodes, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.
+Finally, to enable the simulation of extra-large models with thousands of nodes at billions of samples, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.
 
-\begin{acknowledgements}
-We thank Gautier Stoll for his guidance and fruitful discussion.
-\end{acknowledgements}
+\section{Acknowledgements}
+We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
 
-\begin{funding}
+\section{Funding}
 This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
-\end{funding}
 
-\section*{Bibliography}
+% \section*{Bibliography}
 \bibliography{bibliography}
@@
 %the command above ignores this section for word count
-\onecolumn
-\newpage
+% \onecolumn
+% \newpage
 

### 2024-02-17 mpi-plots/model.R
small plots fix

@@ -1,16 +1,0 @@
-
-x <- read.table('data.tsv', header=T)
-x$size.per.cpu <- x$size/x$cpus
-
-library(dplyr)
-library(tidyr)
-
-x |>
-  group_by(algorithm) |>
-  do(mod = lm(time ~ size.per.cpu + size + cpus, data=., weights=log1p(.$size)),) |>
-  mutate(
-    r.squared = summary(mod)$r.squared,
-    coefs=list(summary(mod)$coefficients[,'Estimate'])
-  ) |>
-  unnest_wider(col=coefs) |>
-  select(-mod) -> result

### 2024-02-17 mpi-plots/plots.R

@@ -1,214 +1,0 @@
-
-if(F) {
-  # use this to install deps
-  install.packages(c('ggplot2', 'cowplot', 'sitools'))
-}
-
-library(ggplot2)
-library(cowplot)
-
-# x <- read.table('data.tsv', header=T)
-
-# #
-# # helpers
-# #
-
-# sival <- function(...) Vectorize(function(x)sitools::f2si(x, ...))
-
-# labeldf <- function(df) {
-#   for(x in names(df))
-#     df[[x]] <- paste(x, '=', df[[x]])
-#   df
-# }
-
-# #
-# # Scaling: how much resources do we need to solve a problem of given expected
-# # size? (is there an unexpected scalability problem?)
-# #
-
-# ggsave("plots/scalability.png", units="in", width=7, height=7,
-# ggplot(x, aes(size, time*cpus/size, color=algorithm)) +
-#   geom_point() +
-#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
-#   scale_x_log10("Base problem size", labels=sival('op')) +
-#   scale_y_log10("Time per operation", labels=sival('Cs/op')) +
-#   ggtitle("Overhead of processing different problem sizes") +
-#   scale_color_brewer("Software", palette='Dark2') +
-#   scale_fill_brewer("Software", palette='Dark2') +
-#   theme_cowplot(font_size=9) +
-#   theme(
-#     panel.grid.major=element_line(size=.2, color='#cccccc'),
-#   )
-# )
-
-# #
-# # Speed: How much does throwing extra CPUs help to solve the problem faster?
-# #
-
-# ggsave("plots/speed.png", units="in", width=7, height=7,
-# ggplot(x, aes(cpus/size, time/size, color=algorithm)) +
-#   geom_point() +
-#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
-#   scale_x_log10("Base operations assigned to one CPU", labels=function(x)(sival('op/C')(1/x))) +
-#   scale_y_log10("Time per base operation", labels=sival('s/op')) +
-#   ggtitle("Speedup given for invested resources") +
-#   scale_color_brewer("Software", palette='Dark2') +
-#   scale_fill_brewer("Software", palette='Dark2') +
-#   theme_cowplot(font_size=9) +
-#   theme(
-#     panel.grid.major=element_line(size=.2, color='#cccccc'),
[…]

### 2024-02-17 Article.tex
giga format
better intro + outro

@@
-\documentclass[a4paper,num-refs]{oup-contemporary}
-% \RequirePackage[2020-02-02]{latexrelease}
-% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
-
-\journal{gigascience}
-
-\usepackage{graphicx}
-\usepackage{siunitx}
-
+\RequirePackage[2020-02-02]{latexrelease}
+\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
 %\usepackage{blindtext}
@@
 
+% Please give the surname of the lead author for the running footer
+\leadauthor{Šmelko}
+
+\begin{document}
+
 \title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
-% \shorttitle{MaBoSS for HPC environments}
+\shorttitle{MaBoSS for HPC environments}
 
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[1,\authfn{1}]{Adam Šmelko}
+\author[1,\Letter]{Adam Šmelko}
 \author[2]{Miroslav Kratochvíl}
 \author[3,4,5]{Emmanuel Barillot}
-% \author[3,4,5]{Laurence Calzone}
-\author[3,4,5,\authfn{1}]{Vincent Noël}
+\author[3,4,5]{Laurence Calzone}
+\author[3,4,5,\Letter]{Vincent Noël}
 
@@
 \affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
-
-\authnote{\authfn{1}smelko@d3s.mff.cuni.cz; vincent.noel@curie.fr}
-
-\papercat{Paper}
-
-\runningauthor{Šmelko et al.}
-
-\jvolume{00}
-\jnumber{0}
-\jyear{2023}
-
-\begin{document}
-
-\begin{frontmatter}
 \maketitle
@@
 \end{keywords}
-\end{frontmatter}
 
-\section{Introduction}
+\begin{corrauthor}
+smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
+\end{corrauthor}
+\section*{Introduction}
 Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer.
@@
 
-\section{Background}
+\section*{Background}
 
@@
 
-\section{Complexity evaluation and parallelization opportunities of MaBoSS}
+\section*{Complexity evaluation and parallelization opportunities of MaBoSS}
 
@@
 
-\subsection{Simulation complexity and parallelization}
+\subsection*{Simulation complexity and parallelization}
 
@@
 
-\subsection{Statistics computation complexity and parallelization}
+\subsection*{Statistics computation complexity and parallelization}
 
@@
 
-\section{Implementation}\label{sec:implementation}
+\section*{Implementation}\label{sec:implementation}
 
@@
 
-\subsection{MaBoSS.GPU}
+\subsection*{MaBoSS.GPU}
 Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective.
@@
 
-\subsubsection{Simulation}
+\subsubsection*{Simulation}
 
@@
 
-\subsubsection{Statistics computation}
+\subsubsection*{Statistics computation}
 
@@
 
-\subsection{MaBoSS.MPI}
+\subsection*{MaBoSS.MPI}
 Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code.
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\subsection{Benchmarking Methodology}
+\subsubsection*{Benchmarking Methodology}
 
@@
 
-\subsection{MaBoSS.GPU}
+\subsection*{MaBoSS.GPU}
 
@@
 
-\subsection{MaBoSS.MPI}
+\subsection*{MaBoSS.MPI}
 
@@
 
-\section{Conclusions}
+\section*{Conclusions}
 
@@
 
-\subsection{Future work}
+\subsection*{Future work}
 
@@
 
-\section{Acknowledgements}
-We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+\begin{acknowledgements}
+We thank Gautier Stoll for his guidance and fruitful discussion.
+\end{acknowledgements}
 
-\section{Funding}
+\begin{funding}
 This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
+\end{funding}
 
-% \section*{Bibliography}
+\section*{Bibliography}
 \bibliography{bibliography}
@@
 %the command above ignores this section for word count
-% \onecolumn
-% \newpage
+\onecolumn
+\newpage
 

### 2024-03-03 Article.tex
a few edits, a few todos

--- 2024-02-14 Article.tex
@@
 \usepackage{algorithm}
-\usepackage{algcompatible}
+\usepackage{algpseudocode}
 
@@
 \author[3,4,5]{Emmanuel Barillot}
-% \author[3,4,5]{Laurence Calzone}
+\author[3,4,5]{Laurence Calzone}
 \author[3,4,5,\authfn{1}]{Vincent Noël}
@@
 \section{Introduction}
-Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer.
-A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
+
+Biological systems are large and complex, and understanding their internal behavior remains critical for designing new therapies for complex diseases such as cancer.
+A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and to predict the efficacy of new treatments.
 Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models.
-While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
-Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.
+While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because of its capability to analyze very large systems.
+Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to advanced properties of cancer~\cite{fumia_carcinogenesis_2013,montagud2022prostate}.
 
 Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
-More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean} to build large Boolean models.
-While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems.
-However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation tools.
-In this effort, adapting existing software to modern HPC computing architectures is an important goal.
+More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
+While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
+However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.
 
-Our work focuses on extending MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
-The main contributions of this paper are twofold:
+Here, we present adaptations of MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains -- to modern HPC computing architectures, which provide significant speedups of the computation, thus allowing scrutinization and analysis of much larger boolean models.
+The main contributions comprise of two new implementations of MaBoSS:
 \begin{itemize}
@@
 
-The source codes of the proposed implementations are publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}.
-We also provide the scripts, presented graphs, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/asmelko/gigascience24-artifact}}.
+The source code of the proposed implementations is publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}.
+We also provide the scripts, presented plots, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/asmelko/gigascience24-artifact}}.
 
-The paper is organized as follows.
-In the next section, we discuss the MaBoSS algorithm and its complexity.
-In the following section, we detail the implementation of MaBoSS.GPU and MaBoSS.MPI.
-In the results section, we present the performance benchmarks of the new implementations.
-Finally, we conclude the paper and discuss the future work.
-[TODO - add section refs]
+To showcase the utility of the new implementations, we performed benchmarking on both existing models and large-scale synthetic models.
+As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speed-ups by utilizing the current HPC infrastructures.
 
@@
 
-We begin with describing the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.
+\subsection{Boolean signaling models}
 
-MaBoSS algorithm takes the Boolean approach for modeling networks which is defined by a Boolean model.
-A model consists of $n$ nodes which can be in one of two Boolean values: active ($1$) or inactive ($0$).
-The \emph{state} of the whole modeled system is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node.
-We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$ and it holds that $|\mathcal{S}| = 2^n$.
+% this is very cool but can we simply refer to the maboss paper? or no one took care to formalize this properly yet?
 
-The model interactions are described as transitions between two states.
-A single state can have multiple transitions to other states; to encode the notion of prioritizing a specific state transition over others, the transitions are assigned a non-negative real value.
-Simply, a Boolean network generalizes to a directed weighted graph $G = (\mathcal{S}, \rho)$, where $\rho: \mathcal{S} \times \mathcal{S} \rightarrow [0, \infty)$ is a transition function generating \emph{transition rates}.
+A Boolean signaling model consists of $n$ nodes which are either active or inactive, gaining values 1 or 0 respectively.
+The \emph{state} of the whole model is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node.
+We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.
+
+Interactions in the model are described as transitions between two states.
+A single state can have multiple transitions to other states with assigned transition probabilities.
+In turn, a Boolean network is represented as a directed weighted graph $G = (\mathcal{S}, \rho)$, where $\rho: \mathcal{S} \times \mathcal{S} \rightarrow [0, \infty)$ is a transition function generating \emph{transition rates}.
 For convenience, it holds that:
-\begin{equation}
-    \rho(S, S') = 0 \iff \text{there is no transition from}\ S\ \text{to}\ S'
-\end{equation}
+\[ \rho(S, S') = 0 \iff \text{there is no transition from}\ S\ \text{to}\ S' \]
+MaBoSS simulates the \emph{asynchronous update strategy}, where at most a single node changes its value in each transition --- consequently, a state $S$ can have at most $n$ possible transitions.
 
-MaBoSS simulates \emph{asynchronous update strategy}, further restricting the transitions between states.
-In this strategy, only a single node can change its value in a single transition; i.e., for a transition from $S$ to $S'$, there exists a single node $i$ such that $S_i \neq S'_i$ and for all $j \neq i$, $S_j = S'_j$.
-Consequently, a state $S$ can have at most $n$ possible transitions.
+To determine the possible transitions, each node follows the \emph{Boolean logic} $\mathcal{B}_i: \mathcal{S} \rightarrow [0, \infty)$, which determines the expected Poisson-process rate of transitioning to the other value.
+If $\mathcal{B}_i(S) = 0$, then the transitions at node $i$ is not allowed in state $S$.
+Given this formalization, the simulation can be also viewed as a continuous-time Markov process.
 
-For each node, a \emph{Boolean logic} $\mathcal{B}: \mathcal{S} \rightarrow [0, \infty)$ is specified.
-The logic of $i$-th node determines if a state $S$ is allowed to transition to such state $S'$, where only $S_i$ and $S'_i$ differ.
-If $\mathcal{B}_i(S) = 0$, then the transition is not allowed.
-Otherwise, the transition is allowed, and the result value determines the transition rate.
+MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$.
+The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a fixed point state with no outgoing transitions.
+The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}.
 
-This formalization can also be understood as a formal definition of the Continuous-time Markov process.
-MaBoSS algorithm simulates this process by producing stochastic \emph{trajectories}.
-A trajectory is a sequence of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$.
-The simulation ends when a state with no outgoing transitions or a maximum simulation time is reached.
+\begin{algorithm}
+\caption{A single iteration of the MaBoSS simulation of a trajectory, given the state $S$ and time $t$.}
+\label{alg:iter}
+\begin{algorithmic}[1]
+\Procedure{TrajectorySimulationStep}{S, t}
+\State $\rho_1 \gets \mathcal{B}_1(S), \dots, \rho_n \gets \mathcal{B}_n(S)$ \Comment compute transition rates
+\State $r \gets \text{random}([0, \sum_{i=1}^n \rho_i))$
+\State $i \gets \min_i \sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$ \Comment select a node for flipping
+\State $S'\gets S\ \text{with the $i$-th bit flipped}$
+\State $u \gets \text{random}([0, 1])$
+\State $\delta t \gets -\frac{\ln u}{\sum_{i=1}^n \rho_i}$
+\State \textbf{return} $(S', t + \delta t)$
+\EndProcedure
+\end{algorithmic}
+%TODO this should be in a float/figure (\begin{algorithm} etc)
+\end{algorithm}
 
-Given a state $S$ and time $t$, the pseudocode for a single iteration of a trajectory simulation is as follows:
-\begin{algorithmic}[1]
-\State Compute transition rates $\rho_1 = \mathcal{B}_1(S), \dots, \rho_n = \mathcal{B}_n(S)$
-\State Sample a random number $r$ from $[0, \sum_{i=1}^n \rho_i)$
-\State Find $i$ such that $\sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$
-\State Compute $S'$ from $S$ by flipping the $i$-th bit
-\State Sample a random number $u$ from $[0, 1]$
-\State Compute $\delta t = -(\ln u) / \sum_{i=1}^n \rho_i$
-\State Set $S = S'$ and $t = t + \delta t$
-\end{algorithmic}
-
-Multiple trajectories are generated to get a good representation of the Markov process. Then, the statistics of the trajectories are computed. These are:
-
-\begin{enumerate}
-    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size $\Delta t$.
+To obtain a good overview of the Markovian process, multiple trajectories are generated and aggregated in a compound trajectory statistics. Commonly obtained statistics include:
+\begin{itemize}
+    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size.
 For each window, the probability of each state is computed as the duration spent in the state divided by the window size.
 The probabilities of the corresponding windows are then averaged across all subtrajectories.
-    \item \emph{Final states} --- The final state of a trajectory is the last state of the trajectory. The sampled states from the trajectories are used to compute the final state distribution.
-    \item \emph{Fixed states} --- A state is fixed if it has no outgoing transitions. The sampled states from the trajectories are used to compute the fixed state distribution.
-\end{enumerate}
-
-Notably, nodes can be marked as \emph{internal}.
-The internal nodes are not considered in the statistics computation of time window probabilities and final states.
-More specifically, before a trajectory is used for statistics computation, the internal nodes are removed from each state in the trajectory.
-This feature is helpful for the models where some nodes are used only for simulation purposes and/or are irrelevant to the final results.
-
-\section{Complexity evaluation and parallelization opportunities of MaBoSS}
-
-MaBoSS algorithm is, in many aspects, very suitable for highly parallel processing.
-The simulation across multiple trajectories is embarrassingly parallel because the trajectories are independent of each other.
-Although the statistics computation is slightly more complicated because the intermediate results from separate trajectories need to be reduced to compute the final statistics, the topic of parallel reduction is a well-researched problem with many efficient solutions.
-
-In this section, we detail the time complexity of the algorithm steps and discuss the parallelization opportunities.
-
-\subsection{Simulation complexity and parallelization}
-
-Let us start by analyzing the time complexity of one iteration step of the simulation.
-Suppose a model with $n$ nodes.
-If we assume that one Boolean logic formula can be evaluated in $\mathcal{O}(n)$, then the computation of transition rates can be done in $\mathcal{O}(n^2)$.
-The selection of the flipping bit is $\mathcal{O}(n)$ and the rest of the iteration is $\mathcal{O}(1)$.
-Therefore, the total time complexity of one iteration is $\mathcal{O}(n^2)$.
-If we simulate $c$ trajectories with an upper bound of trajectory length $u$, the total simulation time is $\mathcal{O}(c \cdot u \cdot n^2)$.
-
-Using an idealized parallel execution machine (PRAM model with infinite parallelism), we can optimize the algorithm in the following ways: 
-\begin{itemize}
-    \item All trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$.
-    \item A single trajectory iteration can be parallelized.
-With $n$ processors, the computation of transition rates is achievable in $\mathcal{O}(n)$ time, where each processor would evaluate one Boolean logic formula, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using parallel prefix sum.
-This sums up to $\mathcal{O}(n)$ time for a single iteration.
+    \item \emph{Final states} --- The last sampled states from the trajectories are used to compute a final state distribution.
+    \item \emph{Fixed states} --- All reached fixed points are used to compute a fixed state distribution.
 \end{itemize}
 
-Overall, using a perfect parallel machine with $c \cdot n$ processors, the simulation time can be reduced to $\mathcal{O}(u \cdot n)$.
-There is a clear dependency between the trajectory iterations, so $\mathcal{O}(u)$ steps must be performed serially.
-Realistically, parallelizing an iteration step may prove beneficial only for rather large models when the scheduling overhead and communications cost are lower than the achieved parallelism.
+To maintain the brevity in the statistics, MaBoSS additionally allows marking some nodes \emph{internal}.
+This is useful because nodes that are not ``interesting'' from the point of final result view occur quite frequently in Boolean models, and removing them from statistics computation often saves a significant amount of resources.
 
-Also, since the trajectory iteration is the main `hot loop' of the whole algorithm (it is executed $c \cdot u$ times), optimizing it as much as possible is imperative.
-The penalty for the processor handling a cache miss inside each loop can result in the orders-of-magnitude slowdown.
-This puts a big pressure on the choice of the data representation for the state $S$ and Boolean logic formulas.
-We discuss this topic in the Implementation section.
+\subsection{Computational complexity of MaBoSS algorithm in parallel}
 
-\subsection{Statistics computation complexity and parallelization}
+\subsubsection{Simulation complexity}
 
-For the sake of brevity, let us discuss only the time window probabilities computation.
-The final and fixed states computation performs the same operation but less often because the final and fixed states are computed only from the last state of each trajectory.
+% TODO this paragraph needs friggin references to the algorithm figure -- took me a while to see where to get the "all other parts"
+We estimate the time required to simulate $c$ trajectories as follows:
+For simplification, we assume that a typical Boolean logic formula in a model of $n$ nodes can be evaluated in $\mathcal{O}(n)$ (this is a very optimistic but empirically valid estimate).
+With that, the computation of all possible transition rates can be finished in $\mathcal{O}(n^2)$.
+The selection of the flipping bit can be finished in $\mathcal{O}(n)$, and all other parts of the iteration can finish in $\mathcal{O}(1)$.
+In total, the time complexity of one iteration is $\mathcal{O}(n^2)$.
+If we simulate $c$ trajectories with an upper bound of trajectory length $u$, the simulation time is in $\mathcal{O}(c \cdot u \cdot n^2)$.
 
-The first part of the computation is updating a data structure that holds the durations of states visited in a trajectory.
-With the choice of a hash map as the data structure, the update can be done in $\mathcal{O}(1)$ time.
-In the second part, after all trajectories are simulated, the intermediate data structures from the trajectories must be merged into a final one.
-This can be done in $\mathcal{O}(\log{c} \cdot m)$ time, assuming a serial merging cost of $\mathcal{O}(m)$.
+In an idealized PRAM model with infinite parallelism, we can optimize the algorithm in the following ways:
+\begin{itemize}
+    \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$.
+    %TODO might be useful to explicitly note that this doesn't include result aggregation (and link to the section)
+    \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
+\end{itemize}
+Thus, using a perfect parallel machine with $c \cdot n$ processors, the computation time can be reduced to $\mathcal{O}(u \cdot n)$.
+Notably, the $\mathcal{O}(u)$ simulation steps that must be performed serially remain a major factor in the whole computation time.
 
-Notably, if the number of states is small enough, or the nodes are dominated by many internal nodes, the hash map can be materialized as a fixed-size array (and be interpreted as a histogram).
-This would eliminate any memory allocations and the key lookup would be reduced to a simple index function.
-This optimization is discussed in the next section.
+\subsubsection{Statistics aggregation}
 
-Finally, we use this discussion to detail the specific optimization choices in the implementation of the algorithms.
+The aggregation of the statistics from the simulations is typically done by updating a shared associative structure indexed by model states, differing only in update frequency between the three kinds of collected statistics.
+
+If the associative structure is implemented as a hashmap, the updates can be done in $\mathcal{O}(1)$ for a single process.
+With multiple processors, the algorithm may hold partial versions of the hashmap for each processor, and aggregate all of them at the end of computation, which can be done in $\mathcal{O}(\log{c} \cdot m)$ using $c$ processors, assuming the maximal size of statistic to be $m$.
+
+As an interesting detail, the hash structures pose a surprising constant-factor overhead.
+In networks where most nodes are internal, the hash map may be replaced by a fixed-size multidimensional array that holds an element for all possible combinations of external node values (basically forming a multidimensional histogram).
+We discuss the impact of this optimization in the next section.
 
@@
 
-In this section, we describe the high-level overview of MaBoSS.GPU and MaBoSS.MPI implementation.
-
 \subsection{MaBoSS.GPU}
-Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective.
-A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware.
-We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.
 
@@
 
-The simulation part is the most computationally demanding part of the CPU MaBoSS implementation.
-The preliminary performance reports show that around $30-80\%$ of the MaBoSS runtime (depending on the model) is spent in the evaluation of Boolean formulas.
-This evaluation is carried out in a recursive call over an expression tree.
-This approach is not well suited for a CPU cache because the whole tree may span multiple cache lines and, apart from other issues, its traversal may easily result in a major cache miss ratio~\cite{karlsson2000prefetching}.
-The GPUs are even more susceptible to this problem because they have a lower cache per core ratio.
-We took this as a starting point for our optimizations.
+In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model).
+The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}
 
-There are multiple ways to tackle the problem of bad cache performance of the expression tree.
-One can either use a linked data structure that is more cache-friendly, such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array with an ad-hoc layout.
-In our approach, we take one step further by utilizing the \emph{runtime compilation of GPU code}~\cite{nvrtc}.
-Using this technique, instead of building the expression tree for the Boolean formulas, we compile the formulas into a native binary code that can be executed directly on the GPU.
-The main advantage of this approach is that the formulas are encoded in the instructions instead of being stored in the memory.
-This way, the formulas would not need to be fetched from memory multiple times during the application runtime.
-The other advantage is that the compiler can optimize the GPU code to utilize the GPU hardware more efficiently.
+There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF bitmasks that can be easily evaluated by vector instructions.
+We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}.
+Using this technique, the Boolean formulas are compiled as functions into a native binary code, which is directly executed by the GPU.
+As the main advantage, the formulae are encoded in the instructions, preventing unnecessary fetches of the encoded formulas from other memory.
+At the same time, the compiler may apply a vast spectrum of optimizations on the Boolean formulas, including case analysis and shortcutting, again resulting in faster evaluation.
 
-The major drawback of the runtime compilation is its speed.
-The compilation takes magnitudes longer than the building of the expression trees.
-On the other hand, the formulas compilation is executed only once at the beginning of the simulation and the compiled code is reused for the whole simulation.
-In theory, the compilation time can get amortized well by the overall simulation and statistics computation time for big enough models.
+A possible drawback of the runtime compilation stems from the relative slowness of the compiler --- for small models, the total execution time of MaBoSS.GPU may be easily dominated by the compilation.
 
-In our implementation, we decided to assign a single GPU thread to a single trajectory.
-We mentioned in the previous section, that a work distribution, where multiple threads would work on a single trajectory, may be beneficial for large models.
-However, this can not be implemented trivially without creating a thread divergence in the transition rates computation (which is usually penalized by a completely serialized execution anyway).
-This optimization requires a clever rewrite of the Boolean logic, which we leave for future work.
+Also, due to the involved implementation complexity, we avoided optimization of the computation of individual trajectories by splitting the Boolean function evaluation to multiple threads (thus missing the factor of $n$ threads from the asymptotic analysis).
+While such optimization might alleviate some cache pressure and thus provide significant performance improvements, we leave its exploration to future work.
 
-\subsubsection{Statistics computation}
+\subsubsection{Statistics aggregation}
 
-The CPU implementation of this part of the algorithm heavily relies on the usage of hash maps.
-They are used to store node states as the keys, and the values may be the state occurrences or their average timespan in the trajectories, depending on the specific statistics computation.
-We will focus on the probability distribution of trajectory states since it is the most computationally demanding part of the statistics computation and the optimizations used for this part apply to the other statistics computations as well.
+For optimizing the statistics aggregation, MaBoSS.GPU heavily relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes, regardless of the size of the model.
+This relatively low number of states generated by non-internal nodes allows us to materialize the whole statistics structure (called ``histogram'') as a fixed-size array (rarely exceeding $2^10$ elements).
 
-The optimization of this step relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes regardless of the total number of nodes.
-The relatively low number of states generated by non-internal nodes allows us to materialize the whole histogram as a fixed-size array.
-The states are no longer stored as the keys --- rather, a hash function is used to map the states to the histogram array indices.
-We use a simple hash function with bit masking and shifting to extract and compact non-internal nodes from the state representation.
-Further, we use well-known GPU histogram optimizations, such as shared memory privatization and atomic operations, to improve the performance of the histogram computation.
+This approach allows us to avoid storing the states as the keys, and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions.
+Further, we use sevelra well-known GPU histogram update optimizations to improve the performance, including as shared memory privatization and atomic operations.
 
 \subsection{MaBoSS.MPI}
-Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code.
-Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory.
-The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.
 
-In summary, MaBoSS.MPI evenly distributes individual trajectories first on MPI nodes and then on the corresponding CPU cores, assuming the MPI nodes are homogenous.
-Once the trajectories are simulated and the statistics are computed for each MPI node, the intermediate data are reduced into a final result using MPI collective operations.
+MaBoSS.MPI is a straightforward extension of the original MaBoSS CPU code to the MPI programming interface.
+Briefly, each thread is assigned to compute a single trajectory at once, progressively collecting the results into a privatized hashmap-based statistics aggregation structure.
+The total amount of trajectory tasks is evenly split among the MPI nodes, adding a second layer of parallelism.
+
+Once all trajectory simulations are finished and the statistics are computed for each thread, the intermediate data are reduced into the final result using MPI collective operations.
 
@@
 
-To evaluate the discussed optimizations, we present the results of performance benchmarks for MaBoSS.GPU and MaBoSS.MPI by comparing their runtimes against the original CPU implementation.
-To get comprehensive results, we used real-world models and synthetic models with varying sizes.
+To evaluate the impact of the implemented optimizations, we present the results of performance benchmarks for MaBoSS.GPU and MaBoSS.MPI by comparing their runtimes against the original CPU implementation.
+To obtain a comprehensive overview of achievable results, we used both real-world models and synthetic models with varying sizes.
 
@@
 
-For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$ (\texttt{cellcycle}~\cite{faure2006cellcycle}, \texttt{sizek}~\cite{sizek2019boolean} and \texttt{Montagud}~\cite{montagud2022prostate}).
-In order to test the scalability of the GPU and MPI implementation, we also created a synthetic models with up to $1000$ nodes.
-Synthetic models are designed in a way such that the length of each trajectory is predictable, and the models have no stable states.
-Also, the number of non-internal nodes is kept low (10 nodes) to enable the usage of the histogram optimization.
+% TODO: ok well, here you have a grande totale insane of 3 models, and decide to omit the size of the middle one by saying "yeah between 10 and 133" :D
+For the benchmarks, we used $3$ real-world models of $10$, $XXX$ and $133$ nodes (\texttt{cellcycle}~\cite{faure2006cellcycle}, \texttt{sizek}~\cite{sizek2019boolean} and \texttt{Montagud}~\cite{montagud2022prostate}).
+In order to test the scalability of the GPU and MPI implementation, we also created several synthetic models with up to $1000$ nodes.
+Synthetic models were designed in a way such that the length of each trajectory is predictable, and the models have no stable states.
+Also, the number of non-internal nodes was kept low (10 nodes) to enable the usage of the histogram optimization.
+%TODO here it would be nice to calm down stupid questions like "ok how realistic are the models actually" by pointing out that they are in the repository or something.
 
@@
 The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2.
-Each measurement was repeated $10$ times, and the average runtime was used for the final results after removing the outliers.
+Each measurement was repeated $10$ times, and the average runtime was used as the final results after removing the outliers.
+%TODO: stats trigger: how did you remove the outliers? Manual QA? What caused the outliers?
 
 The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer.
+% TODO link to MN spec (I think they have a paper)
 
-\subsection{MaBoSS.GPU}
+\subsection{Performance of MaBoSS.GPU}
 
@@
     \includegraphics[width=\linewidth]{Figures/real.pdf}
-    \caption{Wall time comparison of CPU and GPU implementations on real-world models. Each model is simulated with 1 million trajectories.}
+    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on real-world models. Each model is simulated with 1 million trajectories.}
+    % TODO it's slightly inconvenient that the plot legend only describes hardware; in turn the whole thing looks like you're benchmarking the hardware instead of the actual sofware implementations.
+Suggest using: MaBoSS (CPU), MaBoSS.GPU (RTX3070), MaBoSS.GPU (A100).
     \label{fig:real}
@@
 
+% TODO above you're putting the Montagud/Sizek/whatever names into \texttt.
+I'd recommend using \textsc and doing it veeery consistently everywhere.
+Just for the general purpose of sparking the benchmark-nerd imaginations.
 In Figure~\ref{fig:real}, we compare the wall time of the CPU and GPU implementations on real-world datasets.
 The GPU implementation is faster than the CPU implementation on all models, and the speedup shows to be more significant on the models with more nodes and longer trajectories.
-On Montagud model with 133 nodes but a relatively short average trajectory, we achieve $145\times$ speedup.
+On the Montagud model with 133 nodes, but a relatively short average trajectory, we achieve $145\times$ speedup.
 On a slightly smaller Sizek model with a longer average trajectory, the speedup is up to $326\times$.
@@
     \includegraphics[width=\linewidth]{Figures/nodes.pdf}
-    \caption{Wall time comparison of CPU and GPU implementations on synthetic models. Each model is simulated with 1 million trajectories.}
+    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models.
+Each model is simulated with 1 million trajectories.
+The two panels differ by inclusion of the runtime GPU code compilation, showing its impact on total run time.}
     \label{fig:synth}
@@
 Figure~\ref{fig:synth} shows much finer performance progression on synthetic models.
-On the left subplot, the runtime compilation is filtered out to show the pure simulation performance.
-We can see that the CPU variant starts to progress steeper at around the $100$ nodes boundary.
-We can expect that the implementation hits the cache size limit and fetching the required data from the memory comes with more overhead.
-The same can be observed in the GPU variant, but at $200$ nodes and with more drastic increase.
-This observation corresponds to a much more significant performance penalty when fetching data from the GPU memory.
-Overall, this suggests that the optimization of dividing transition rates computation among multiple threads mentioned in the previous section may provide a significant speedup for bigger models as it alleviates the register pressure.
+We observed that the CPU variant starts to progress steeper at around the $100$ nodes boundary.
+We assume that the implementation hits the cache size limit, and the overhead of fetching the required data from the memory becomes dominant.
+The same can be observed in the GPU variant later at around $200$ nodes.
+Expectably, the cache-spilling performance penalty is much more significant on GPUs.
+Overall, the results suggest that the optimization of dividing transition rate computations among multiple threads, as mentioned in the previous section XXX TODO REF XXX, may provide a better speedup for bigger models, as it alleviates the register and cache pressure.
+% TODO danger there's a non-commented TODO above. Add an actual \ref if possible.
 
@@
     \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
+    % TODO: what happens if we swap the sample count (color) with formula length (facet) ?
+    % Also, it might be much better to have the percentages as logarithmic scale. (most of the space is now killed by the single 200node 10formula point)
+    % TODO: "sample count" in the legend is not defined anywhere around, maybe say "total trajectories" ?
     \label{fig:comp}
@@
 
-The right subplot shows the total runtime of the GPU implementation.
-Comparing the subplots shows that the runtime compilation ratio decreases with increasing model size.
-Figure~\ref{fig:comp} offers more detailed benchmarks for this scenario, run on the NVIDIA Tesla A100 GPU.
-The compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).
-Notably, as soon as the simulation becomes more complex, i.e., by increasing the number of nodes or simulated trajectories, the compilation time becomes less than $10\%$ of the total runtime, even for the models with unrealistically long formulae.
-This suggests that the runtime compilation is a viable optimization for bigger models.
+Additionally, Figure~\ref{fig:synt} shows the total runtime of the GPU implementation including the runtime compilation step.
+Comparing the panels, we observe that that the relative runtime compilation overhead quickly disappears with increasing model size.
+Figure~\ref{fig:comp} shows the results of a more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU.
+We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).
+Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae.
+This suggests that the runtime compilation is a viable optimization methodology also for much larger models.
 
-\subsection{MaBoSS.MPI}
+\subsection{Performance of MaBoSS.MPI}
+
+%TODO achtung, the figure DPI is somehow messed up and the letters are too tiny.
+%TODO: might be better to merge the figures into one (2 facets)
+%TODO: it's nice that there's a trend on the diagonal but that kinda looks like we're hiding fluctuations or small off-angle, can we scale the thing into a visibly horizontal straight line?
+%  - multiply the walltime by the number of cores to produce the realistic totalwastedtime
+%  - divide the speedup by the number of cores invested (to show speedup per investment)
 
@@
 
-Figure~\ref{fig:sizek_results} shows the efficiency of the MPI implementation on the Sizek model.
+Figure~\ref{fig:sizek_results} shows the efficiency of the MaBoSS.MPI implementation on the Sizek model.
 We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores.
 We can observe a close-to-linear speedup of up to 64 MPI nodes (1280 cores), and a plateau for larger suites (Figure~\ref{fig:sizek_results}, green).
-This can be explained by hitting a bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.
+This can be explained by hitting an expectable bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.
 
@@
 
-To stress the scalability of the implementation, we also used a synthetic model with 1000 nodes (see Figure~\ref{fig:synthetic_results}).
+To stress the scalability of the implementation, we also used a synthetic model with 1000 nodes.
 We simulated this model on 32 cores per MPI node, on 1 to 192 nodes (32 to 6144 cores).
-With this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes.
-As we hypothesized, a decline in the speedup was not observed in larger simulations.
+The obtained speedups are summarized in Figure~\ref{fig:synthetic_results}.
+Using this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes.
+As expected, the plateau in the speedup was not observed in simulations that involve larger models.
 
@@
 
-In this work, we presented two new implementations of MaBoSS tool, a continuous time Boolean model simulator, designed for HPC environments:  MaBoSS.GPU which is designed to exploit the computational power of massively parallel GPU hardware, and MaBoSS.MPI which enables MaBoSS to scale to many nodes of HPC clusters.
+In this work, we presented two new implementations of MaBoSS tool, a continuous time Boolean model simulator, both of which are designed to enable utilization of the HPC computing resources: MaBoSS.GPU is designed to exploit the computational power of massively parallel GPU hardware, and MaBoSS.MPI enables MaBoSS to scale to many nodes of HPC clusters via the MPI framework.
 We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both variants are capable of providing significant speedups over the original CPU code.
-The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.
+The GPU implementation shows 145--326\texttimes\ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.
 
-Overall, we believe the new MaBoSS extensions will enable the simulation of large, automatically generated models and will be a valuable tool for the systems biology community.
+Overall, we believe that the new MaBoSS implementations enable simulation and exploration of the behavior of very large, automatically generated models, thus becoming a valuable analysis tool for the systems biology community.
 
@@
 
-As mentioned in the discussion section, the next steps in the development of MaBoSS.GPU would be to parallelize the iteration step of the simulation.
-This requires a different representation of Boolean formulas, which would not pose any thread divergence, probably by substituting runtime compilation with an in-memory data structure.
-This optimization could potentially decrease the register pressure created by holding the state data and thus increase the performance of huge models with hundreds of nodes.
+During the development, we identified several optimization directions that could be taken by researchers to further scale up the MaBoSS simulation approach.
 
-Furthermore, many parallelization methods used in GPU implementation could be backported to the CPU implementation, increasing the overall usability of MaBoSS running on systems without GPU accelerators.
+Mainly, the parallelization scheme used in MaBoSS.GPU could be enhanced to also parallelize over the evaluation of Boolean formulas.
+To avoid GPU thread divergence, this would however require a specialized Boolean formula representation, entirely different from the current version of MaBoSS; likely even denying the relative efficiency of the use of runtime compilation.
+On the other hand, this optimization might decrease the register pressure created by holding the state data, and thus increase the performance on models with thousands of nodes.
 
-Finally, to enable the simulation of extra-large models with thousands of nodes at billions of samples, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.
+In the long term, easier optimization paths might lead to sufficiently good results: For example, backporting the GPU implementation improvements back to the MaBoSS CPU implementation could improve the performance even on systems where GPU accelerators are not available.
+Similarly, both MaBoSS.GPU and MaBoSS.MPI could be combined into a single software that executes distributed GPU-based analysis over multiple MPI nodes, giving a single high-performance solution for extremely large problems.
 
 \section{Acknowledgements}
-We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+We thank Gautier Stoll for his guidance and fruitful discussion.
 
 \section{Funding}
-This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
+%TODO the permed ack wasn't right, it should be:
+he research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
+% This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]
+The project was partially supported by Charles University, SVV project number~260698.
 

### 2024-03-03 Article.tex
a few edits, a few todos
plot some plotty things

@@
 \author[3,4,5]{Emmanuel Barillot}
-\author[3,4,5]{Laurence Calzone}
+% \author[3,4,5]{Laurence Calzone}
 \author[3,4,5,\authfn{1}]{Vincent Noël}
@@
 \end{algorithmic}
-%TODO this should be in a float/figure (\begin{algorithm} etc)
 \end{algorithm}
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/sizek_mpi.pdf}
+\includegraphics[width=3.25in]{Figures/sizek_mpi.pdf}
 \caption{Scalability results of MPI implementation on Sizek model with 20 cores per MPI node.}
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/synth_mpi_speedup.pdf}
+\includegraphics[width=3.25in]{Figures/synth_mpi_speedup.pdf}
 \caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and 32 cores per MPI node.}
@@
 \section{Acknowledgements}
-We thank Gautier Stoll for his guidance and fruitful discussion.
+We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
 

### 2024-03-03 mpi-plots/plots.R

@@ -1,0 +1,211 @@
+
+if(F) {
+  # use this to install deps
+  install.packages(c('ggplot2', 'cowplot', 'sitools'))
+}
+
+library(ggplot2)
+library(cowplot)
+
+# x <- read.table('data.tsv', header=T)
+
+# #
+# # helpers
+# #
+
+# sival <- function(...) Vectorize(function(x)sitools::f2si(x, ...))
+
+# labeldf <- function(df) {
+#   for(x in names(df))
+#     df[[x]] <- paste(x, '=', df[[x]])
+#   df
+# }
+
+# #
+# # Scaling: how much resources do we need to solve a problem of given expected
+# # size? (is there an unexpected scalability problem?)
+# #
+
+# ggsave("plots/scalability.png", units="in", width=7, height=7,
+# ggplot(x, aes(size, time*cpus/size, color=algorithm)) +
+#   geom_point() +
+#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
+#   scale_x_log10("Base problem size", labels=sival('op')) +
+#   scale_y_log10("Time per operation", labels=sival('Cs/op')) +
+#   ggtitle("Overhead of processing different problem sizes") +
+#   scale_color_brewer("Software", palette='Dark2') +
+#   scale_fill_brewer("Software", palette='Dark2') +
+#   theme_cowplot(font_size=9) +
+#   theme(
+#     panel.grid.major=element_line(size=.2, color='#cccccc'),
+#   )
+# )
+
+# #
+# # Speed: How much does throwing extra CPUs help to solve the problem faster?
+# #
+
+# ggsave("plots/speed.png", units="in", width=7, height=7,
+# ggplot(x, aes(cpus/size, time/size, color=algorithm)) +
+#   geom_point() +
+#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
+#   scale_x_log10("Base operations assigned to one CPU", labels=function(x)(sival('op/C')(1/x))) +
+#   scale_y_log10("Time per base operation", labels=sival('s/op')) +
+#   ggtitle("Speedup given for invested resources") +
+#   scale_color_brewer("Software", palette='Dark2') +
+#   scale_fill_brewer("Software", palette='Dark2') +
+#   theme_cowplot(font_size=9) +
+#   theme(
[…]

### 2024-03-08 Article.tex
author change
major revision

@@
 To showcase the utility of the new implementations, we performed benchmarking on both existing models and large-scale synthetic models.
-As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speed-ups by utilizing the current HPC infrastructures.
+As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speedups by utilizing the current HPC infrastructures.
 
@@
 
-\subsection{Computational complexity of MaBoSS algorithm in parallel}
+\subsection{Computational complexity of parallel MaBoSS algorithm}
 
@@
 
-% TODO this paragraph needs friggin references to the algorithm figure -- took me a while to see where to get the "all other parts"
 We estimate the time required to simulate $c$ trajectories as follows:
 For simplification, we assume that a typical Boolean logic formula in a model of $n$ nodes can be evaluated in $\mathcal{O}(n)$ (this is a very optimistic but empirically valid estimate).
-With that, the computation of all possible transition rates can be finished in $\mathcal{O}(n^2)$.
-The selection of the flipping bit can be finished in $\mathcal{O}(n)$, and all other parts of the iteration can finish in $\mathcal{O}(1)$.
+With that, the computation of all possible transition rates (Algorithm~\ref{alg:iter}, line $2$) can be finished in $\mathcal{O}(n^2)$.
+The selection of the flipping bit (Algorithm~\ref{alg:iter}, line $4$) can be finished in $\mathcal{O}(n)$, and all other parts of the iteration can finish in $\mathcal{O}(1)$.
 In total, the time complexity of one iteration is $\mathcal{O}(n^2)$.
@@
     \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$.
-    %TODO might be useful to explicitly note that this doesn't include result aggregation (and link to the section)
+(Note that this does not include the results aggregation.
+See \emph{Statistics aggregation} section for further description.)
     \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
@@
 In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model).
-The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}
+The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.
 
@@
 For optimizing the statistics aggregation, MaBoSS.GPU heavily relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes, regardless of the size of the model.
-This relatively low number of states generated by non-internal nodes allows us to materialize the whole statistics structure (called ``histogram'') as a fixed-size array (rarely exceeding $2^10$ elements).
+This relatively low number of states generated by non-internal nodes allows us to materialize the whole statistics structure (called ``histogram'') as a fixed-size array (rarely exceeding $2^{10}$ elements).
 
 This approach allows us to avoid storing the states as the keys, and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions.
-Further, we use sevelra well-known GPU histogram update optimizations to improve the performance, including as shared memory privatization and atomic operations.
+Further, we use several well-known GPU histogram update optimizations to improve the performance, including shared memory privatization and atomic operations.
 
@@
 
-% TODO: ok well, here you have a grande totale insane of 3 models, and decide to omit the size of the middle one by saying "yeah between 10 and 133" :D
-For the benchmarks, we used $3$ real-world models of $10$, $XXX$ and $133$ nodes (\texttt{cellcycle}~\cite{faure2006cellcycle}, \texttt{sizek}~\cite{sizek2019boolean} and \texttt{Montagud}~\cite{montagud2022prostate}).
+For the benchmarks, we used $3$ real-world models of $10$, $87$ and $133$ nodes (\textsc{cellcycle}~\cite{faure2006cellcycle}, \textsc{sizek}~\cite{sizek2019boolean} and \textsc{Montagud}~\cite{montagud2022prostate}).
 In order to test the scalability of the GPU and MPI implementation, we also created several synthetic models with up to $1000$ nodes.
@@
 Also, the number of non-internal nodes was kept low (10 nodes) to enable the usage of the histogram optimization.
-%TODO here it would be nice to calm down stupid questions like "ok how realistic are the models actually" by pointing out that they are in the repository or something.
+The synthetic models together with their python generator are available in the replication package.
 
@@
 The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2.
-Each measurement was repeated $10$ times, and the average runtime was used as the final results after removing the outliers.
-%TODO: stats trigger: how did you remove the outliers? Manual QA? What caused the outliers?
+Each measurement was repeated $10$ times, and the average runtime was used as the final results.
 
-The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer.
-% TODO link to MN spec (I think they have a paper)
+The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer~\footnote{\url{https://www.bsc.es/marenostrum/marenostrum}}.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{Figures/real.pdf}
+    \includegraphics[width=\linewidth]{plots/real.pdf}
     \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on real-world models. Each model is simulated with 1 million trajectories.}
-    % TODO it's slightly inconvenient that the plot legend only describes hardware; in turn the whole thing looks like you're benchmarking the hardware instead of the actual sofware implementations.
-Suggest using: MaBoSS (CPU), MaBoSS.GPU (RTX3070), MaBoSS.GPU (A100).
     \label{fig:real}
@@
 
-% TODO above you're putting the Montagud/Sizek/whatever names into \texttt.
-I'd recommend using \textsc and doing it veeery consistently everywhere.
-Just for the general purpose of sparking the benchmark-nerd imaginations.
 In Figure~\ref{fig:real}, we compare the wall time of the CPU and GPU implementations on real-world datasets.
 The GPU implementation is faster than the CPU implementation on all models, and the speedup shows to be more significant on the models with more nodes and longer trajectories.
-On the Montagud model with 133 nodes, but a relatively short average trajectory, we achieve $145\times$ speedup.
-On a slightly smaller Sizek model with a longer average trajectory, the speedup is up to $326\times$.
+On the \textsc{Montagud} model with 133 nodes, but a relatively short average trajectory, we achieve $145\times$ speedup.
+On a slightly smaller \textsc{sizek} model with a longer average trajectory, the speedup is up to $326\times$.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{Figures/nodes.pdf}
+    \includegraphics[width=\linewidth]{plots/nodes.pdf}
     \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models.
@@
 Expectably, the cache-spilling performance penalty is much more significant on GPUs.
-Overall, the results suggest that the optimization of dividing transition rate computations among multiple threads, as mentioned in the previous section XXX TODO REF XXX, may provide a better speedup for bigger models, as it alleviates the register and cache pressure.
-% TODO danger there's a non-commented TODO above. Add an actual \ref if possible.
+Overall, the results suggest that the optimization of dividing transition rate computations among multiple threads, as mentioned in \emph{Implementation} section, may provide a better speedup for bigger models, as it alleviates the register and cache pressure.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{Figures/nodes-compilation-big-NVIDIA Tesla A100 GPU.pdf}
+    \includegraphics[width=\linewidth]{plots/nodes-compilation-big-NVIDIA Tesla A100 (MaBoSS.GPU).pdf}
     \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
-    % TODO: what happens if we swap the sample count (color) with formula length (facet) ?
-    % Also, it might be much better to have the percentages as logarithmic scale. (most of the space is now killed by the single 200node 10formula point)
-    % TODO: "sample count" in the legend is not defined anywhere around, maybe say "total trajectories" ?
     \label{fig:comp}
@@
 
-Additionally, Figure~\ref{fig:synt} shows the total runtime of the GPU implementation including the runtime compilation step.
+Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step.
 Comparing the panels, we observe that that the relative runtime compilation overhead quickly disappears with increasing model size.
@@
 
-%TODO achtung, the figure DPI is somehow messed up and the letters are too tiny.
-%TODO: might be better to merge the figures into one (2 facets)
-%TODO: it's nice that there's a trend on the diagonal but that kinda looks like we're hiding fluctuations or small off-angle, can we scale the thing into a visibly horizontal straight line?
-%  - multiply the walltime by the number of cores to produce the realistic totalwastedtime
-%  - divide the speedup by the number of cores invested (to show speedup per investment)
-
 \begin{figure}%[tbhp]
 \centering
-\includegraphics[width=3.25in]{Figures/sizek_mpi.pdf}
+\includegraphics[width=3.25in]{plots/sizek_mpi.pdf}
 \caption{Scalability results of MPI implementation on Sizek model with 20 cores per MPI node.}
@@
 
-Figure~\ref{fig:sizek_results} shows the efficiency of the MaBoSS.MPI implementation on the Sizek model.
+Figure~\ref{fig:sizek_results} shows the efficiency of the MaBoSS.MPI implementation on the \textsc{sizek} model.
 We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores.
@@
 \centering
-\includegraphics[width=3.25in]{Figures/synth_mpi_speedup.pdf}
+\includegraphics[width=3.25in]{plots/synth_mpi_speedup.pdf}
 \caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and 32 cores per MPI node.}
@@
 % \centering
-% \includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
+% \includegraphics[width=.8\linewidth]{plots/large_model.pdf}
 % \caption{Scalability results for synthetic model with 1000 nodes}
@@
 %\centering
-%\includegraphics[width=.8\linewidth]{Figures/Figure_1}
+%\includegraphics[width=.8\linewidth]{plots/Figure_1}
 %\caption{Placeholder image of Iris with a long example caption to show justification setting.}
@@
 \section{Funding}
-%TODO the permed ack wasn't right, it should be:
-he research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
+The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
 % This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]

### 2024-03-08 mpi-plots/plots.R

@@ -1,211 +1,0 @@
-
-if(F) {
-  # use this to install deps
-  install.packages(c('ggplot2', 'cowplot', 'sitools'))
-}
-
-library(ggplot2)
-library(cowplot)
-
-# x <- read.table('data.tsv', header=T)
-
-# #
-# # helpers
-# #
-
-# sival <- function(...) Vectorize(function(x)sitools::f2si(x, ...))
-
-# labeldf <- function(df) {
-#   for(x in names(df))
-#     df[[x]] <- paste(x, '=', df[[x]])
-#   df
-# }
-
-# #
-# # Scaling: how much resources do we need to solve a problem of given expected
-# # size? (is there an unexpected scalability problem?)
-# #
-
-# ggsave("plots/scalability.png", units="in", width=7, height=7,
-# ggplot(x, aes(size, time*cpus/size, color=algorithm)) +
-#   geom_point() +
-#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
-#   scale_x_log10("Base problem size", labels=sival('op')) +
-#   scale_y_log10("Time per operation", labels=sival('Cs/op')) +
-#   ggtitle("Overhead of processing different problem sizes") +
-#   scale_color_brewer("Software", palette='Dark2') +
-#   scale_fill_brewer("Software", palette='Dark2') +
-#   theme_cowplot(font_size=9) +
-#   theme(
-#     panel.grid.major=element_line(size=.2, color='#cccccc'),
-#   )
-# )
-
-# #
-# # Speed: How much does throwing extra CPUs help to solve the problem faster?
-# #
-
-# ggsave("plots/speed.png", units="in", width=7, height=7,
-# ggplot(x, aes(cpus/size, time/size, color=algorithm)) +
-#   geom_point() +
-#   geom_smooth(method='loess', aes(fill=algorithm), alpha=.1) +
-#   scale_x_log10("Base operations assigned to one CPU", labels=function(x)(sival('op/C')(1/x))) +
-#   scale_y_log10("Time per base operation", labels=sival('s/op')) +
-#   ggtitle("Speedup given for invested resources") +
-#   scale_color_brewer("Software", palette='Dark2') +
-#   scale_fill_brewer("Software", palette='Dark2') +
-#   theme_cowplot(font_size=9) +
-#   theme(
[…]

### 2024-03-08 plots.R

@@ -1,0 +1,221 @@
+library(ggplot2)
+library(cowplot)
+library(sitools)
+library(dplyr)
+
+
+W=4.804
+H=2
+S=1
+point_size=0.8
+line_size=0.5
+linecolors=scale_color_brewer(palette="Set2")
+theme = theme_cowplot(font_size=7)
+
+sisec=Vectorize(function(t)if(is.na(t))NA else sitools::f2si(t / 10^6, 's'))
+
+average <- function(data)
+{
+    data["total"] = data["compilation"] + data["visualization"] + data["simulation"]
+    data <- subset(data, select = -X)
+
+    data_s <- split(data, paste0(data$max_time, data$nodes, data$formula_size, data$sample_count))
+    data <- NULL
+    for (i in 1:length(data_s)){
+        tmp = subset(data_s[[i]], !(total %in% boxplot(data_s[[i]]$total, plot = FALSE)$out))
+        data <- rbind(data, tmp)
+    }
+
+    data = data.frame(data %>% group_by_at(names(data)[-grep("(total)|(compilation)|(visualization)|(simulation)", names(data))]) %>% summarise(total = mean(total), compilation = mean(compilation), visualization = mean(visualization), simulation = mean(simulation)))
+}
+
+data_gpu = read.csv('data/gpu_out.csv', header=T, sep=';')
+data_gpu = average(data_gpu)
+
+data_v100 = read.csv('data/gpu_out_v100.csv', header=T, sep=';')
+data_v100 = average(data_v100)
+
+data_a100 = read.csv('data/gpu_out_a100.csv', header=T, sep=';')
+data_a100 = average(data_a100)
+
+data_cpu = read.csv('data/cpu_out.csv', header=T, sep=';')
+names(data_cpu)[names(data_cpu) == "parsing"] <- "compilation"
+data_cpu = subset(data_cpu, select = -threads)
+data_cpu = average(data_cpu)
+
+data_cpu["type"] = "Intel Xeon Gold 6130 (MaBoSS.CPU)"
+data_gpu["type"] = "NVIDIA RTX 3070 Laptop (MaBoSS.GPU)"
+data_v100["type"] = "NVIDIA Tesla V100 (MaBoSS.GPU)"
+data_a100["type"] = "NVIDIA Tesla A100 (MaBoSS.GPU)"
+data = rbind(data_cpu, data_gpu, data_a100)
+
+
+data_gpu_real = read.csv('data/gpu_out_real.csv', header=T, sep=';')
[…]

### 2024-03-17 Article.tex
bioarxiv template

@@
-\documentclass[a4paper,num-refs]{oup-contemporary}
-% \RequirePackage[2020-02-02]{latexrelease}
-% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
-
-\journal{gigascience}
+\RequirePackage[2020-02-02]{latexrelease}
+\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
 
@@
 
+% Please give the surname of the lead author for the running footer
+\leadauthor{Šmelko}
+
+\begin{document}
+
 \title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
-% \shorttitle{MaBoSS for HPC environments}
+\shorttitle{MaBoSS for HPC environments}
 
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[1,\authfn{1}]{Adam Šmelko}
+\author[1,\Letter]{Adam Šmelko}
 \author[2]{Miroslav Kratochvíl}
@@
 % \author[3,4,5]{Laurence Calzone}
-\author[3,4,5,\authfn{1}]{Vincent Noël}
+\author[3,4,5,\Letter]{Vincent Noël}
 
@@
 \end{keywords}
-\end{frontmatter}
 
-\section{Introduction}
+\begin{corrauthor}
+    smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
+    \end{corrauthor}
+
+\section*{Introduction}
 
@@
 Here, we present adaptations of MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains -- to modern HPC computing architectures, which provide significant speedups of the computation, thus allowing scrutinization and analysis of much larger boolean models.
-The main contributions comprise of two new implementations of MaBoSS:
+The main contributions comprise two new implementations of MaBoSS:
 \begin{itemize}
@@
 To showcase the utility of the new implementations, we performed benchmarking on both existing models and large-scale synthetic models.
-As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speedups by utilizing the current HPC infrastructures.
+As the main results, MaBoSS.GPU provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speedups by utilizing the current HPC infrastructures.
 
-\section{Background}
+\section*{Background}
 
-\subsection{Boolean signaling models}
+\subsection*{Boolean signaling models}
 
@@
 To determine the possible transitions, each node follows the \emph{Boolean logic} $\mathcal{B}_i: \mathcal{S} \rightarrow [0, \infty)$, which determines the expected Poisson-process rate of transitioning to the other value.
-If $\mathcal{B}_i(S) = 0$, then the transitions at node $i$ is not allowed in state $S$.
+If $\mathcal{B}_i(S) = 0$, then the transition at node $i$ is not allowed in state $S$.
 Given this formalization, the simulation can be also viewed as a continuous-time Markov process.
@@
 
-To obtain a good overview of the Markovian process, multiple trajectories are generated and aggregated in a compound trajectory statistics. Commonly obtained statistics include:
+To obtain a good overview of the Markovian process, multiple trajectories are generated and aggregated in compound trajectory statistics. Commonly obtained statistics include:
 \begin{itemize}
@@
 
-\subsection{Computational complexity of parallel MaBoSS algorithm}
+\subsection*{Computational complexity of parallel MaBoSS algorithm}
 
-\subsubsection{Simulation complexity}
+\subsubsection*{Simulation complexity}
 
@@
 
-\subsubsection{Statistics aggregation}
+\subsubsection*{Statistics aggregation}
 
@@
 If the associative structure is implemented as a hashmap, the updates can be done in $\mathcal{O}(1)$ for a single process.
-With multiple processors, the algorithm may hold partial versions of the hashmap for each processor, and aggregate all of them at the end of computation, which can be done in $\mathcal{O}(\log{c} \cdot m)$ using $c$ processors, assuming the maximal size of statistic to be $m$.
+With multiple processors, the algorithm may hold partial versions of the hashmap for each processor, and aggregate all of them at the end of the computation, which can be done in $\mathcal{O}(\log{c} \cdot m)$ using $c$ processors, assuming the maximal size of statistic to be $m$.
 
@@
 
-\section{Implementation}\label{sec:implementation}
+\section*{Implementation}\label{sec:implementation}
 
-\subsection{MaBoSS.GPU}
+\subsection*{MaBoSS.GPU}
 
-\subsubsection{Simulation}
+\subsubsection*{Simulation}
 
 In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model).
-The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.
+The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issues) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.
 
@@
 
-Also, due to the involved implementation complexity, we avoided optimization of the computation of individual trajectories by splitting the Boolean function evaluation to multiple threads (thus missing the factor of $n$ threads from the asymptotic analysis).
+Also, due to the involved implementation complexity, we avoided optimization of the computation of individual trajectories by splitting the Boolean function evaluation into multiple threads (thus missing the factor of $n$ threads from the asymptotic analysis).
 While such optimization might alleviate some cache pressure and thus provide significant performance improvements, we leave its exploration to future work.
 
-\subsubsection{Statistics aggregation}
+\subsubsection*{Statistics aggregation}
 
@@
 
-\subsection{Benchmarking Methodology}
+\subsection*{Benchmarking Methodology}
 
@@
 Also, the number of non-internal nodes was kept low (10 nodes) to enable the usage of the histogram optimization.
-The synthetic models together with their python generator are available in the replication package.
+The synthetic models together with their Python generator are available in the replication package.
 
@@
 The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2.
-Each measurement was repeated $10$ times, and the average runtime was used as the final results.
+Each measurement was repeated $10$ times, and the average runtime was used as the final result.
 
@@
 
-\subsection{Performance of MaBoSS.GPU}
+\subsection*{Performance of MaBoSS.GPU}
 
@@
 Each model is simulated with 1 million trajectories.
-The two panels differ by inclusion of the runtime GPU code compilation, showing its impact on total run time.}
+The two panels differ by the inclusion of the runtime GPU code compilation, showing its impact on total run time.}
     \label{fig:synth}
@@
 Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step.
-Comparing the panels, we observe that that the relative runtime compilation overhead quickly disappears with increasing model size.
-Figure~\ref{fig:comp} shows the results of a more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU.
+Comparing the panels, we observe that the relative runtime compilation overhead quickly disappears with increasing model size.
+Figure~\ref{fig:comp} shows the results of more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU.
 We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).
@@
 
-\section{Acknowledgements}
+\begin{acknowledgements}
 We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+\end{acknowledgements}
 
-\section{Funding}
+\begin{funding}
 The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
@@
 The project was partially supported by Charles University, SVV project number~260698.
+\end{funding}
 
-% \section*{Bibliography}
+\section*{Bibliography}
 \bibliography{bibliography}

### 2024-03-17 Article.tex
small fixes

@@
-\RequirePackage[2020-02-02]{latexrelease}
-\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
+\documentclass[a4paper,num-refs]{oup-contemporary}
+% \RequirePackage[2020-02-02]{latexrelease}
+% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
+
+\journal{gigascience}
 
@@
 
-% Please give the surname of the lead author for the running footer
-\leadauthor{Šmelko}
-
-\begin{document}
-
 \title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
-\shorttitle{MaBoSS for HPC environments}
+% \shorttitle{MaBoSS for HPC environments}
 
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[1,\Letter]{Adam Šmelko}
+\author[1,\authfn{1}]{Adam Šmelko}
 \author[2]{Miroslav Kratochvíl}
@@
 % \author[3,4,5]{Laurence Calzone}
-\author[3,4,5,\Letter]{Vincent Noël}
+\author[3,4,5,\authfn{1}]{Vincent Noël}
 
@@
 \end{keywords}
+\end{frontmatter}
 
-\begin{corrauthor}
-    smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
-    \end{corrauthor}
-
-\section*{Introduction}
+\section{Introduction}
 
@@
 
-\section*{Background}
+\section{Background}
 
-\subsection*{Boolean signaling models}
+\subsection{Boolean signaling models}
 
@@
 
-\subsection*{Computational complexity of parallel MaBoSS algorithm}
+\subsection{Computational complexity of parallel MaBoSS algorithm}
 
-\subsubsection*{Simulation complexity}
+\subsubsection{Simulation complexity}
 
@@
 
-\subsubsection*{Statistics aggregation}
+\subsubsection{Statistics aggregation}
 
@@
 
-\section*{Implementation}\label{sec:implementation}
+\section{Implementation}\label{sec:implementation}
 
-\subsection*{MaBoSS.GPU}
+\subsection{MaBoSS.GPU}
 
-\subsubsection*{Simulation}
+\subsubsection{Simulation}
 
@@
 
-\subsection*{Benchmarking Methodology}
+\subsection{Benchmarking Methodology}
 
@@
 
-\subsection*{Performance of MaBoSS.GPU}
+\subsection{Performance of MaBoSS.GPU}
 
@@
 
-\section*{Conclusions}
+\section{Conclusions}
 
@@
 
-\begin{acknowledgements}
+\section{Acknowledgements}
 We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
-\end{acknowledgements}
 
-\begin{funding}
+\section{Funding}
 The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
@@
 The project was partially supported by Charles University, SVV project number~260698.
-\end{funding}
 
-\section*{Bibliography}
+% \section*{Bibliography}
 \bibliography{bibliography}

### 2024-03-22 Article.tex
update

@@
 
-\papercat{Paper}
+\papercat{Technical Note}
 
@@
 The source code of the proposed implementations is publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}.
-We also provide the scripts, presented plots, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/asmelko/gigascience24-artifact}}.
+We also provide the scripts, presented plots, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/sysbio-curie/gigascience24-artifact}}.
 
@@
 
-\section{Acknowledgements}
-We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
+\section{Availability of source code and requirements}
+% flex and bison are linux only :/
+\begin{itemize}
+	\item Project name: MaBoSS.GPU
+	\item Project home page: \url{https://github.com/sysbio-curie/MaBoSS.GPU}
+	\item Operating system(s): Platform independent
+	\item Programming language: C++, CUDA
+	\item Other requirements: Flex, Bison, CMake >= 3.18, Cuda toolkit >= 12.0
+	\item License: MIT
+\end{itemize}
+
+\begin{itemize}
+	\item Project name: MaBoSS.MPI
+	\item Project home page: \url{https://github.com/sysbio-curie/MaBoSS}
+	\item Operating system(s): Platform independent
+	\item Programming language: C++
+	\item Other requirements: Flex, Bison
+	\item License: BSD3-clause
+\end{itemize}
+
+\section{Data availability}
+The scripts, presented plots, data and instructions to reproduce the benchmarks are available on GitHub~\cite{adam_smelko_2024_10841007}.
+
+\subsection{Competing Interests}
+The authors declare that they have no competing interests.
 
@@
 The project was partially supported by Charles University, SVV project number~260698.
+
+\section{Acknowledgements}
+We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
 

### 2024-04-25 rebuttal.txt
some gpu related fixes

First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks. We have addressed them in our paper revision. The individual answers are in the following. Minor remarks that were simply fixed in the paper and required no additional comments are omitted.

Reviewer #1:

Q: How is the parallelization for GPUs actually done? In the section MaBoSS.GPU, it is described, why there was no parallelization over the Boolean functions, but not about the rest. In the MPI section, it states, that each thread is assigned to compute a single trajectory at once. I would assume that this is also true for the GPU implementation, this should, however, be made clear.
A: You are right, we included this piece of information into the last paragraph of Simulation section.

Q: The legend in the figures needs to be more specific to be able to interpret the figures better: e.g. how many synthetic models? (are the 19 points the respective models?)
A: yes, the x-axis shows multiple synthetic models of various sizes. We added this to Figure 2.

Q: In Figure1, the High-End NVIDIA TESLA is either slower or as fast as the consumer grade laptop GPU. Is this purely due to the compilation time (as Figure 2 would suggest, that it makes a difference)? 
A: You are right. You can see for yourself by comparing gpu_out_real.csv and gpu_out_real_a100.csv files in the artifact. We are unsure why the compilation takes longer for high-end GPU than for laptop GPU, after all, it is a proprietary piece of software. Perhaps the loading of shared libraries takes a different amount of time for the operating systems... Due to the uncertainty and the fact that the difference is in hundreds of milliseconds, we do not see this as a detail worth explicitly mentioning in the paper.

Q: There is no demonstration of how a user would actually deal with the code - especially since the GPU version is based on a hard-coded implementation of the rules. A demonstrative example would help to get an idea, how a user would actually get started, and how deep he needs to dig into the code - especially for a biological audience, accessability of such tools is an important point
A: Respectfully, it is not true that the GPU version is based on any hard-coded implementation. The compilation of the rules happens during the runtime without any user intervention. We included this note in the second paragraph of the Simulation section.

Reviewer #2:

Q: Figure 3: The x-axis states “Nodes count (log−scale)”, however, it appears to have a linear scale. The linear relationships claimed in the statement “We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).” is not clear from the provided figures, and should be clarified or, for example provide Figure 3 with a linear-linear or log-log scale.
A: Thanks for pointing out the wrong scale in the plot, it is a linear scale. Regarding the linear dependency, it can not be seen from Fig 3 since it shows compilation time percentages rather than the absolute values. Therefore, plotting a linear-linear scale would not show any linear dependency. Nevertheless, the stated linear dependency is indeed there and we observed it from the plots which we did not include in the paper. The general idea why it holds is that increasing the number of nodes by a value of x adds O(x) new lines to a source file which needs to be compiled. We included this explanation in the paper.

### 2024-04-26 rebuttal.txt
some minor

@@
+- Extend literature in the intro (R1 Q1)
+- More detailed background 
+-- Add a new background subsection describing CPU impl (R1 Q2,3,7)
+-- Elaborate more on MaBoSS output (perhaps a new subsection with plotted outputs) (R1 Q4,7 R2 Q1)
+-- Revise the algorithm (R1 Q5,6)
+- Revise GPU stuff (R1 Q8,10,11,13 R2 Q2,3.1,3.3,3.5)
+- Revise MPI stuff (R2 Q3.2,3.4)
+- in Results add a table with models we used ? (R1 Q9)
+- explain the MPI outlier (remeasure?) (R1 Q12) 
+- minor stuff (rest)
+
 First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks.
@@
 I would assume that this is also true for the GPU implementation, this should, however, be made clear.
-A: You are right, we included this piece of information into the last paragraph of Simulation section.
+A: You are right, we included this piece of information in the last paragraph of Simulation section.
 
@@
 Perhaps the loading of shared libraries takes a different amount of time for the operating systems...
-Due to the uncertainty and the fact that the difference is in hundreds of milliseconds, we do not see this as a detail worth explicitly mentioning in the paper.
+We added a note in the Performance of MaBoSS.GPU section.
 
@@
 We included this explanation in the paper.
+
+Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
+A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a fix.

### 2024-04-26 plots.R
more minor stuff

@@ -121,7 +121,7 @@ for (m in c("NVIDIA RTX 3070 Laptop (MaBoSS.GPU)", "NVIDIA Tesla A100 (MaBoSS.GP
         ggplot(data_c, aes(x=nodes,y=ratio, color=sample_count, shape=sample_count)) +
         geom_point(size=point_size) +
         geom_line(linewidth=line_size) +
-        xlab("Nodes count (log-scale)")+
+        xlab("Nodes count")+
         ylab("% of total time compiling (log-scale)")+
         labs(color="Simulated Trajectories", shape="Simulated Trajectories") +
         #scale_color_manual(values=RColorBrewer::brewer.pal(9,'YlGnBu')[2:9]) +

### 2024-04-26 plots.R
increase font of mpi plots

@@ -174,7 +174,7 @@ data <- subset(data, algorithm %in% c('mpi', 'new'))
 
 data$algorithm <- factor(data$algorithm, levels=c('mpi', 'new'), labels=c('MPI', 'CPU'))
 
-ggsave("plots/sizek_mpi.pdf", units="in", width=3.5, height=2,
+ggsave("plots/sizek_mpi.pdf", units="in", width=4, height=2,
 ggplot(data, aes(cpus, time, color=algorithm, shape=algorithm, group=algorithm)) +
   geom_point() +
   stat_smooth(geom='line', method='loess') +
@@ -182,7 +182,7 @@ ggplot(data, aes(cpus, time, color=algorithm, shape=algorithm, group=algorithm))
   scale_y_log10("Wall time (log-scale)", labels=c(expression('10'^1*' s'),expression('10'^2*' s'),expression('10'^3*' s'),expression('10'^4*' s')), breaks=c(10,100,1000,10000)) +
   scale_color_brewer("Software", palette='Dark2') +
   scale_shape("Software") +
-  theme_cowplot(font_size=5) +
+  theme_cowplot(font_size=9) +
   theme(
     panel.grid.major=element_line(size=.2, color='#cccccc'),
   )
@@ -192,14 +192,14 @@ data <- read.table('data/mpi_synth.tsv', header=T)
 
 data$algorithm <- factor(data$algorithm, levels=c('mpi'), labels=c('MPI'))
 
-ggsave("plots/synth_mpi.pdf", units="in", width=3.25, height=2,
+ggsave("plots/synth_mpi.pdf", units="in", width=4, height=2,
 ggplot(data, aes(cpus, time*cpus, color=algorithm, shape=algorithm, group=algorithm)) +
   geom_point() +
   scale_x_log10("Cores (log-scale)") +
   scale_y_log10("Total wall time (all CPUs, log-scale)", labels=c(expression('1 x 10'^3*' s'),expression('3 x 10'^3*' s'),expression('1 x 10'^4*' s'),expression('3 x 10'^4*' s')), breaks=c(1000,3000,10000,30000)) +
   scale_color_brewer("Software", palette='Dark2') +
   scale_shape("Software") +
-  theme_cowplot(font_size=5) +
+  theme_cowplot(font_size=9) +
   theme(
     panel.grid.major=element_line(size=.2, color='#cccccc'),
   )
[…]

### 2024-04-30 rebuttal.txt
add model table

@@
 Q: The legend in the figures needs to be more specific to be able to interpret the figures better: e.g. how many synthetic models? (are the 19 points the respective models?)
-A: yes, the x-axis shows multiple synthetic models of various sizes. We added this to Figure 2.
+A: Yes, the x-axis shows multiple synthetic models of various sizes. We added this to Figure 2.
 
@@
 
+Q: It is stated that synthetic models with up to 1000 nodes were created.
+The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective).
+A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.
+
 Reviewer #2:
@@
 Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
-A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a fix.
+A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a workaround.
+
+Q: The MaBoSS.GPU README could do with a link to documentation.
+A: We added a new section to README with a high-level architectural overview.

### 2024-05-01 rebuttal.txt
maboss output section

@@
 
+Q: The statistics are described, but how would a user interpret them/make use of them for themselves?
+What is this shared associative structure that the model is updating?
+Maybe a diagram would be helpful for visualization?
+A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.
+
+Q: How does the code actually deal with circular steady states?
+A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops.
+Their statistics are further computed without any change.
+The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper.
+Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.
+
 Reviewer #2:
@@
 A: We added a new section to README with a high-level architectural overview.
+
+Q: Since the authors have disabled issue tracking on their github repositories, I urge them to add a contact method to sysbio-curie/hpcmaboss-artifact and sysbio-curie/MaBoSS.GPU, for users to obtain assistance and provide feedback, including bug reports.
+A: This was a mistake, we now have enabled issue tracking on MaBoSS.GPU and hpcmaboss-artifact repositories. Thanks for reminding us.
+
+Q:  It is not clear whether the version provided via the colomoto conda repository is MPI enabled, so a clarification is needed.
+A: This is something we have been considering, but we have not moved forward yet.
+We will most likely have to create specific conda packages which are MPI enabled, in addition to the classic ones.
+Same thing for MaBoSS.GPU.
+We also need to think about this for the colomoto docker, but again it might not be straightforward.
+For now, we think that the target users of these new implementations are knowledgeable enough to install them manually.
+
+Q: A concern is that the provided background is insufficient for the applicability and usefulness of the software to be apparent to the broader audience of BMC Bioinformatics.
+The authors should add more to provide context and insight into the interpretation and use of MaBoSS output, for those not already familiar with the previous MaBoSS versions.
+A: We included a new section Statistics output and visualization.

### 2024-05-02 rebuttal.txt
cpu version rebuttal

@@
 
+Q: There needs to be a description about the original code.
+Most importantly, it is always referred to, as the CPU version.
+This suggests, that we are talking about a purely sequential code.
+In the Methodology, however, it is described to be run on a full 32-core node, which implies already some basic parallelization.
+This should be made clear from the beginning.
+A: Indeed, the implementation of the "CPU version" was early on parallelized to work on multiple cores, using POSIX threads.
+We added a new subsection to the background section (MaBoSS CPU Implementation) to describe this implementation.
+
 Reviewer #2:

### 2024-05-03 rebuttal.txt
finish background

@@
 
+Q: The Boolean signaling modeling paragraph needs to be extended in more details: To me it is not clear, why e.g., the transition rates should be in [0, inf) and not [0,1], since we are talking about probabilities.
+The concept of asynchronous updating should be made a bit clearer as well.
+Furthermore, it is often referred to the evaluation of the Boolean formulas as the most expensive computational part.
+It would help to get an idea, how such a formula would look like, and how they are implemented, to give the reader context about this problem.
+Furthermore, the word "node" is used quite often in the paper, so it should be made clear, what this refers to exactly.
+A: We extended the background section with all your comments.
+Regarding the node, [0, inf) transition rates and asynchronous updates, we added explanations to the first paragraphs of sections 2.1 and 2.2.
+In order to add an example of Boolean formula, we explained it in more detail by defining its binary function, which determines the transition rate using a simple rule in equation 5.
+Having the binary function, we presented its simple example in equation 4.
+The way how the boolean formulas are implemented in the code is described in the first paragraph of section 3.1, simply referring to it as a recursive traversal of an expression tree.
+
+Q: Algorithm 1 should be made clearer: how are transition rates computed? What does the bit-flip mean? delta t is computed using a "u" - where does that come from?
+A: Transition rates are computed by evaluating boolean formulas of each node, as is written on line 2 of Algorithm 1.
+To make it clearer, we modified the second paragraph of section 2.2, where we explicitly state, that the output of boolean logic is the transition rates.
+Regarding "u", the Gillespie algorithm uses 2 random numbers (r,u) drawn from the uniform distribution to randomly select the transition node and the transition time.
+We added comments to the algorithm and added a line referencing that the algorithm follows the Gillespie algorithm.
+
 Reviewer #2:
@@
 A: We included a new section Statistics output and visualization.
+
+Q: Finally, to improve accessibility, the following could do with explanation: variables “u” and “δt” in Algorithm 1, “PRAM”, “CNF”, “DNF”.
+A: We explained the acronyms and provided comments for “δt” in the algorithm.
+“u” is to provide randomness in “δt” selection, which is now highlighted in the comment.
+Moreover, we reference the Gillespie algorithm, on which our algorithm is based.

### 2024-05-05 rebuttal.txt
explain outlier

@@
 
+Q: Please ellaborate more on the last point in Figure 5, that seems to be an outlier. Which synthetic model is this? Is it also found in the earlier figures? 
+A: Firstly, we included additional information about the synthetic model in the figure caption.
+Secondly, we analyzed the datapoint and it is the same case as in the previous MPI figure - the problem is distributed into so many small tasks that the actual cost of MPI communication becomes the bottleneck.
+We are sorry that we did not include this information in the first version - initially the figure was plotted in a different way which did not show this drastic speedup difference, after replotting the datapoint went unnoticed.
+
 Reviewer #2:

### 2024-05-05 rebuttal.txt
reorder rebuttal

--- 2024-05-03 rebuttal.txt
@@
-- Extend literature in the intro (R1 Q1)
-- More detailed background 
--- Add a new background subsection describing CPU impl (R1 Q2,3,7)
--- Elaborate more on MaBoSS output (perhaps a new subsection with plotted outputs) (R1 Q4,7 R2 Q1)
--- Revise the algorithm (R1 Q5,6)
-- Revise GPU stuff (R1 Q8,10,11,13 R2 Q2,3.1,3.3,3.5)
-- Revise MPI stuff (R2 Q3.2,3.4)
-- in Results add a table with models we used ? (R1 Q9)
-- explain the MPI outlier (remeasure?) (R1 Q12) 
-- minor stuff (rest)
-
 First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks.
@@
 
+Q: There needs to be a description about the original code.
+Most importantly, it is always referred to, as the CPU version.
+This suggests, that we are talking about a purely sequential code.
+In the Methodology, however, it is described to be run on a full 32-core node, which implies already some basic parallelization.
+This should be made clear from the beginning.
+A: Indeed, the implementation of the "CPU version" was early on parallelized to work on multiple cores, using POSIX threads.
+We added a new subsection to the background section (MaBoSS CPU Implementation) to describe this implementation.
+
+Q: How does the code actually deal with circular steady states?
+A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops.
+Their statistics are further computed without any change.
+The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper.
+Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.
+
+Q: The Boolean signaling modeling paragraph needs to be extended in more details: To me it is not clear, why e.g., the transition rates should be in [0, inf) and not [0,1], since we are talking about probabilities.
+The concept of asynchronous updating should be made a bit clearer as well.
+Furthermore, it is often referred to the evaluation of the Boolean formulas as the most expensive computational part.
+It would help to get an idea, how such a formula would look like, and how they are implemented, to give the reader context about this problem.
+Furthermore, the word "node" is used quite often in the paper, so it should be made clear, what this refers to exactly.
+A: We extended the background section with all your comments.
+Regarding the node, [0, inf) transition rates and asynchronous updates, we added explanations to the first paragraphs of sections 2.1 and 2.2.
+In order to add an example of Boolean formula, we explained it in more detail by defining its binary function, which determines the transition rate using a simple rule in equation 5.
+Having the binary function, we presented its simple example in equation 4.
+The way how the boolean formulas are implemented in the code is described in the first paragraph of section 3.1, simply referring to it as a recursive traversal of an expression tree.
+We believe that for the sake of giving the context to boolean logic evaluation, we included enough information.
+If the reader wants to dive into details, we extended the readme in maboss.GPU repository with a detailed description of formulas implementation.
+
+Q: Algorithm 1 should be made clearer: how are transition rates computed? What does the bit-flip mean? delta t is computed using a "u" - where does that come from?
+A: Transition rates are computed by evaluating boolean formulas of each node, as is written on line 2 of Algorithm 1.
+To make it clearer, we modified the second paragraph of section 2.2, where we explicitly state, that the output of boolean logic is the transition rates.
+Regarding "u", the Gillespie algorithm uses 2 random numbers (r,u) drawn from the uniform distribution to randomly select the transition node and the transition time.
+We added comments to the algorithm and added a line referencing that the algorithm follows the Gillespie algorithm.
+
+Q: The statistics are described, but how would a user interpret them/make use of them for themselves?
+What is this shared associative structure that the model is updating?
+Maybe a diagram would be helpful for visualization?
+A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.
+
 Q: How is the parallelization for GPUs actually done?
@@
 A: You are right, we included this piece of information in the last paragraph of Simulation section.
+
+Q: It is stated that synthetic models with up to 1000 nodes were created.
+The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective).
+A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.
 
@@
 
+Q: Please ellaborate more on the last point in Figure 5, that seems to be an outlier. Which synthetic model is this? Is it also found in the earlier figures? 
+A: Firstly, we included additional information about the synthetic model in the figure caption.
+Secondly, we analyzed the datapoint, and it is the same case as in the previous MPI figure - the problem is distributed into so many small tasks that the actual cost of MPI communication becomes the bottleneck.
+We are sorry that we did not include this information in the first version - initially, the figure was plotted in a different way, which did not show this drastic speedup difference; after replotting, the datapoint went unnoticed.
+
 Q: There is no demonstration of how a user would actually deal with the code - especially since the GPU version is based on a hard-coded implementation of the rules.
@@
 
-Q: It is stated that synthetic models with up to 1000 nodes were created.
-The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective).
-A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.
+Reviewer #2:
 
-Q: The statistics are described, but how would a user interpret them/make use of them for themselves?
-What is this shared associative structure that the model is updating?
-Maybe a diagram would be helpful for visualization?
-A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.
-
-Q: How does the code actually deal with circular steady states?
-A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops.
-Their statistics are further computed without any change.
-The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper.
-Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.
-
-Q: There needs to be a description about the original code.
-Most importantly, it is always referred to, as the CPU version.
-This suggests, that we are talking about a purely sequential code.
-In the Methodology, however, it is described to be run on a full 32-core node, which implies already some basic parallelization.
-This should be made clear from the beginning.
-A: Indeed, the implementation of the "CPU version" was early on parallelized to work on multiple cores, using POSIX threads.
-We added a new subsection to the background section (MaBoSS CPU Implementation) to describe this implementation.
-
-Q: The Boolean signaling modeling paragraph needs to be extended in more details: To me it is not clear, why e.g., the transition rates should be in [0, inf) and not [0,1], since we are talking about probabilities.
-The concept of asynchronous updating should be made a bit clearer as well.
-Furthermore, it is often referred to the evaluation of the Boolean formulas as the most expensive computational part.
-It would help to get an idea, how such a formula would look like, and how they are implemented, to give the reader context about this problem.
-Furthermore, the word "node" is used quite often in the paper, so it should be made clear, what this refers to exactly.
-A: We extended the background section with all your comments.
-Regarding the node, [0, inf) transition rates and asynchronous updates, we added explanations to the first paragraphs of sections 2.1 and 2.2.
-In order to add an example of Boolean formula, we explained it in more detail by defining its binary function, which determines the transition rate using a simple rule in equation 5.
-Having the binary function, we presented its simple example in equation 4.
-The way how the boolean formulas are implemented in the code is described in the first paragraph of section 3.1, simply referring to it as a recursive traversal of an expression tree.
-
-Q: Algorithm 1 should be made clearer: how are transition rates computed? What does the bit-flip mean? delta t is computed using a "u" - where does that come from?
-A: Transition rates are computed by evaluating boolean formulas of each node, as is written on line 2 of Algorithm 1.
-To make it clearer, we modified the second paragraph of section 2.2, where we explicitly state, that the output of boolean logic is the transition rates.
-Regarding "u", the Gillespie algorithm uses 2 random numbers (r,u) drawn from the uniform distribution to randomly select the transition node and the transition time.
-We added comments to the algorithm and added a line referencing that the algorithm follows the Gillespie algorithm.
-
-Reviewer #2:
+Q: A concern is that the provided background is insufficient for the applicability and usefulness of the software to be apparent to the broader audience of BMC Bioinformatics.
+The authors should add more to provide context and insight into the interpretation and use of MaBoSS output, for those not already familiar with the previous MaBoSS versions.
+A: We included a new section Statistics output and visualization.
 
@@
 We included this explanation in the paper.
-
-Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
-A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a workaround.
 
@@
 
-Q: A concern is that the provided background is insufficient for the applicability and usefulness of the software to be apparent to the broader audience of BMC Bioinformatics.
-The authors should add more to provide context and insight into the interpretation and use of MaBoSS output, for those not already familiar with the previous MaBoSS versions.
-A: We included a new section Statistics output and visualization.
+Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
+A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a workaround.
 
